### 2020-05-05 main.Rmd
methods 1

---
title             : "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
shorttitle        : "CNNs for classifying effector proteins"

author: 
  - name          : "Ruth Kristianingsih"
    affiliation   : "1"
    email         : "ruth.kristianingsih@tsl.ac.uk"
  - name          : "Dan MacLean"
    email         : "dan.maclean@tsl.ac.uk"
    address       : "The Sainsbury Laboratory, University of East Anglia, Norwich, UK, NR4 7UH"
    corresponding : yes    # Define only one corresponding author
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "The Sainsbury Laboratory"

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.

  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.

  One sentence clearly stating the **general problem** being addressed by  this particular study.

  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).

  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.

  One or two sentences to put the results into a more **general context**.

  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.

  <!-- https://tinyurl.com/ybremelq -->

  <!-- keywords          : "keywords" 
  wordcount         : "X" -->

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Methods

## Sequence Data Collection

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the team, the file can be accessed at ^[Ruth please provide this file, is there a link in a repo?]. The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps

```{r, effectorcollection, fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to Uniprot"}
knitr::include_graphics(
  "figures/effectorsequence_collection.png",
                        )
```

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results

## Sequence Collection

The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that as high quality set of annotated effectors as possible was collected. To this end we used PHI-Base [@phibase] as our primary sequence store. Sequences in PHI-Base are human curated from the literature and have therefore been noted in experimental studies. They do not derive from large scale annotations or contain hypothetical or predicted proteins. This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifially reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal. To collect effector sequences we parsed a whole database text dump of version 4.8 ^[Ruth, who provided the text dump, was it Martin Urban? Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and Uniprot IDs, then the sequence retrieved from UNIPROT. The sequences retrieved can be seen in the data file in this manuscript's repository ('effector_information.txt') ^[Ruth, can you point out where this data file is in your development repository]. Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly. The total number of plant effectors per group can be seen in Table \@ref(tab:effsummary). The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable)

Sequences for non-effector, secreted proteins were collected using a similar pipeline. Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool. For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected an identical number of non-effector, secreted proteins to that collected in the effector set. This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples ^[Ruth, can you add more detail]

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Effectors found in three major pathogen groups'
)
```

```{r, bacefftable}
apa_table(
  readr::read_tsv("tables/bacterial_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
)
```
```{r, fungiefftable}
apa_table(
  readr::read_tsv("tables/fungal_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Fungal species contributing to the fungal effector sequence set retrieved from PHI-Base'
)
```
```{r, oomefftable}
apa_table(
  readr::read_tsv("tables/oomycete_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Oomycete species contributing to the oomycete effector sequence set retrieved from PHI-Base'
)
```

# Discussion

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2020-05-05 main.Rmd
method 1

@@
 Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the team, the file can be accessed at ^[Ruth please provide this file, is there a link in a repo?].
-The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps
+The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used.
+We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release ^[Ruth, can you confirm the UniProt version you accessed and data], using the code in ^[Ruth, can you add the script/workbook URL from your repo here, please].
+We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn.
+We downloaded sequences for randomly selected proteins matching these criteria from  Ensembl databases ^[Ruth, please list the Ensemble data bases you use] using the BioMart tools with the code in ^[Ruth, can you add the script/workbook URL from your repo here, please].
 
-```{r, effectorcollection, fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to Uniprot"}
+```{r, effectorcollection, fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
 knitr::include_graphics(
@@
 To collect effector sequences we parsed a whole database text dump of version 4.8 ^[Ruth, who provided the text dump, was it Martin Urban?
-Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and Uniprot IDs, then the sequence retrieved from UNIPROT.
+Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and UniProt IDs, then the sequence retrieved from UniProt.
 The sequences retrieved can be seen in the data file in this manuscript's repository ('effector_information.txt') ^[Ruth, can you point out where this data file is in your development repository].

### 2020-05-05 main.Rmd
updated main.Rmd

@@
 
-Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the team, the file can be accessed at ^[Ruth please provide this file, is there a link in a repo?].
+Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the team, the file can be accessed at [github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
 The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used.
-We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release ^[Ruth, can you confirm the UniProt version you accessed and data], using the code in ^[Ruth, can you add the script/workbook URL from your repo here, please].
+We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release `2019_05`, using the code in [github.com/TeamMacLean/ruth-effectors-prediction/](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd).
 We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn.
-We downloaded sequences for randomly selected proteins matching these criteria from  Ensembl databases ^[Ruth, please list the Ensemble data bases you use] using the BioMart tools with the code in ^[Ruth, can you add the script/workbook URL from your repo here, please].
+We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [Ensembl Fungi](https://fungi.ensembl.org/index.html) and [Ensembl Protists](https://protists.ensembl.org/index.html) manually using the BioMart tools.
+Since BioMart tool is not supported the Ensembl Bacteria, then we downloaded whole genome data matching the bacteria effector data from [Ensembl Bacteria](http://bacteria.ensembl.org/index.html) using the code in [github.com/TeamMacLean/ruth-effectors-prediction](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
+Additionally, we used SignalP 3.0 in order to get the secreted sequences.
 
@@
 This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifially reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal.
-To collect effector sequences we parsed a whole database text dump of version 4.8 ^[Ruth, who provided the text dump, was it Martin Urban?
-Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and UniProt IDs, then the sequence retrieved from UniProt.
-The sequences retrieved can be seen in the data file in this manuscript's repository ('effector_information.txt') ^[Ruth, can you point out where this data file is in your development repository].
+To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv) ^[Ruth, who provided the text dump, was it Martin Urban?
+Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and UniProt IDs.
+These ID then were mapped to PHI-Base fasta.
+If the sequences are not available in PHI-Base fasta data, then the sequence retrieved from UniProt.
+The sequences retrieved can be seen in the data file in this manuscript's repository [(`effector_data.csv`)](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv).
 Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly.
@@
 For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected an identical number of non-effector, secreted proteins to that collected in the effector set.
-This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples ^[Ruth, can you add more detail]
+This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples.
 

### 2020-05-05 main.Rmd
added flowchart on the Rmd

@@
 
-```{r, effectorcollection, fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
 knitr::include_graphics(
-  "figures/effectorsequence_collection.png",
+  "figures/flowchart_effectorsequence_collection.pdf",
+)
+```
+
+```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+knitr::include_graphics(
+  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
 )

### 2020-05-06 main.Rmd
updated the description on getting secreted non-effector data

@@
 We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [Ensembl Fungi](https://fungi.ensembl.org/index.html) and [Ensembl Protists](https://protists.ensembl.org/index.html) manually using the BioMart tools.
-Since BioMart tool is not supported the Ensembl Bacteria, then we downloaded whole genome data matching the bacteria effector data from [Ensembl Bacteria](http://bacteria.ensembl.org/index.html) using the code in [github.com/TeamMacLean/ruth-effectors-prediction](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
+Since BioMart tool is not supported the Ensembl Bacteria, then we downloaded whole genome data matching the bacteria effector data from [Ensembl Bacteria FTP](https://bacteria.ensembl.org/info/website/ftp/index.html) using the code in [github.com/TeamMacLean/ruth-effectors-prediction](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
 Additionally, we used SignalP 3.0 in order to get the secreted sequences.
@@
 This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples.
+Figure \@ref(fig:noneffectorcollection) summaries the process of getting non-effector set.
+Using the gene information obtained from effector data, we downloaded secreted sequences from the same genome from each individual effector sequence depending on the availability of them in Ensembl Fungi and Ensembl Protists, otherwise we downloaded sequences from the same species as the individual effector using BioMart tool.
+However, BioMart tool is not supported for Ensembl Bacteria.
+Therefore, we downloaded the bacteria sequences using Ensmbl Bacteria FTP.
+Additionaly, we only selected the secreted sequences from prediction results of SignalP ran on the downloaded data.
+We then blast the combination of both secreted sequences data downloaded manually using BioMart tool and using Ensembl FTP with the effector sequence data, and removed identical sequences.
+This process is done in order to make sure there is no effector data in our candidate non-effector sample.
+Last step in this process then was to sample an identical number of non-effector data as in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable).
+In this step, we also blast every time we took a sample, so that there is no identical sequences within non-effector set.
 

### 2020-05-06 main.Rmd
corrected minor mistake in spelling and added the path of non-effector data

@@
 This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifially reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal.
-To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv) ^[Ruth, who provided the text dump, was it Martin Urban?
-Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and UniProt IDs.
+To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered with IDs and UniProt IDs.
 These ID then were mapped to PHI-Base fasta.
@@
 Figure \@ref(fig:noneffectorcollection) summaries the process of getting non-effector set.
-Using the gene information obtained from effector data, we downloaded secreted sequences from the same genome from each individual effector sequence depending on the availability of them in Ensembl Fungi and Ensembl Protists, otherwise we downloaded sequences from the same species as the individual effector using BioMart tool.
+Using the gene information obtained from effector data, we downloaded secreted sequences from the same genome from each individual effector sequence depending on their availability on Ensembl Fungi and Ensembl Protists, otherwise we downloaded sequences from the same species as the individual effector.
+BioMart tool is used in this process.
 However, BioMart tool is not supported for Ensembl Bacteria.
-Therefore, we downloaded the bacteria sequences using Ensmbl Bacteria FTP.
+Therefore, we downloaded the bacteria sequences using Ensembl Bacteria FTP.
 Additionaly, we only selected the secreted sequences from prediction results of SignalP ran on the downloaded data.
@@
 Last step in this process then was to sample an identical number of non-effector data as in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable).
-In this step, we also blast every time we took a sample, so that there is no identical sequences within non-effector set.
+In this step, we also ran blast every time we took a sample, so that there is no identical sequences within non-effector set.
+The non-effector set sequences obtained can be seen in the data file in this manuscript's repository [(`non_effector_data.csv`)](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).
 

### 2020-05-11 main.Rmd
table of hyperparams and method of hyperparam

--- 2020-05-05 main.Rmd
@@
 library("papaja")
+library(magrittr)
 ```
@@
 
-Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the team, the file can be accessed at [github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
+Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
 The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used.
-We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release `2019_05`, using the code in [github.com/TeamMacLean/ruth-effectors-prediction/](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd).
+We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd).
 We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn.
-We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [Ensembl Fungi](https://fungi.ensembl.org/index.html) and [Ensembl Protists](https://protists.ensembl.org/index.html) manually using the BioMart tools.
-Since BioMart tool is not supported the Ensembl Bacteria, then we downloaded whole genome data matching the bacteria effector data from [Ensembl Bacteria](http://bacteria.ensembl.org/index.html) using the code in [github.com/TeamMacLean/ruth-effectors-prediction](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
-Additionally, we used SignalP 3.0 in order to get the secreted sequences.
+We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates_ensembl_2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley_biomart_2009].
+Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP.
+With these we used SignalP 3.0 [@dyrlov_bendtsen_improved_2004] in order to filter the secreted sequences and selected accordingly.
+^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camacho_blast_2009].
+We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
 
@@
 
+## Encoding and Subsetting Sequences
+
+The sequences collected were encoded using either one-hot encoding (CNN-LSTM based models) or integer based encoding (CNN-GRU-LSTM models).
+Encoded sequences were split into taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd)
+
+## Model Training
+
+We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding.
+We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R ^[Ruth, Version?], Tensorflow ^[Ruth, Version?], Keras ^[Ruth, Version?], Python ^[Ruth, Version?] on machines ^[Ruth, Machine types] with  Gb RAM ^[Ruth, Amounts?].
+
+### Hyperparameter Scans
+
+We performed a random search of hyperparameter effects by selecting ^[Ruth, how many?
+The report says 50, but it isn't clear whether this is 50 from 100 possible, or 50 from 50 possible - when you write something you need to keep in mind the question - 'How would my reader repeat this?' ] of all possible hyperparameters for each model type at random ^[Ruth, how is this done?
+Is there a script to reference?].
+We applied five-fold cross validation at each combination and recorded model performance on the validation data at each iteration.
+Code for this can be found ^[Ruth, where?].
+Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam)
+
+### Fine tuning
+
+With best performing models from hyperparameter scans we carried out a process of fine tuning manually to ensure that models did not over- or underfit data and were trained to be as general as possible.
+We loaded best performing models and assessed changes in selected parameters as described in Results to produce final models ^[Ruth, did you use anything like tensorboard or just R to manually tune the models?].
+Once each model was tuned accordingly using the training and test sets, final classification testing and estimation of accuracy was done with the held-out validation set that the model had not seen until this point.
+The final models were saved into HDF5 objects and stored in the repository at [link](link) ^[Ruth, link please!]
+
 ## Procedure
@@
 
-The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that as high quality set of annotated effectors as possible was collected.
-To this end we used PHI-Base [@phibase] as our primary sequence store.
+The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that we collected as high a quality set of annotated effectors as possible.
+To this end we used PHI-Base [@phibase] as our primary sequence origin.
 Sequences in PHI-Base are human curated from the literature and have therefore been noted in experimental studies.
 They do not derive from large scale annotations or contain hypothetical or predicted proteins.
-This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifially reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal.
-To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv) ^[Ruth, who provided the text dump, was it Martin Urban?
-Where is the file in your repos?], all proteins marked as plant pathogen effectors were filtered with IDs and UniProt IDs.
-These ID then were mapped to PHI-Base fasta.
-If the sequences are not available in PHI-Base fasta data, then the sequence retrieved from UniProt.
-The sequences retrieved can be seen in the data file in this manuscript's repository [(`effector_data.csv`)](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv).
+This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifically reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal.
+To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID.
+The sequences and IDs retrieved can be seen in the data file in this manuscript's repository [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv).
 Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly.
@@
 Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool.
-For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected an identical number of non-effector, secreted proteins to that collected in the effector set.
+For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected an identical number of non-effector, secreted proteins to that collected in the effector set from either the same strain or species.
 This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples.
+Figure \@ref(fig:noneffectorcollection) summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).
+
+## Model Training
+
+TODO:
+
+comes to fig  21, 22, 23
 
@@
 
+```{r, hyperparam }
+ readr::read_tsv("tables/hyperparamscan_info.tab") %>% 
+  knitr::kable( "latex", longtable = T, booktabs = T,
+           digits = 0,
+  align = 'c',
+  caption = 'Hyperparameters used in automatic scanning of hyperparameter space and best performing parameters values for each model'
+         ) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
+
+```
 # Discussion

### 2020-05-12 main.Rmd
completed the link of repos and the details of software used for the projects

@@
 We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding.
-We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R ^[Ruth, Version?], Tensorflow ^[Ruth, Version?], Keras ^[Ruth, Version?], Python ^[Ruth, Version?] on machines ^[Ruth, Machine types] with  Gb RAM ^[Ruth, Amounts?].
+We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM.
 
@@
 
-We performed a random search of hyperparameter effects by selecting ^[Ruth, how many?
-The report says 50, but it isn't clear whether this is 50 from 100 possible, or 50 from 50 possible - when you write something you need to keep in mind the question - 'How would my reader repeat this?' ] of all possible hyperparameters for each model type at random ^[Ruth, how is this done?
-Is there a script to reference?].
+We performed a random search of hyperparameter effects by selecting 50 of all possible hyperparameters depending on each model type at random.
+The total of possible hyperparameter combination for each model can be found in Table \@ref(tab:possiblecombination).
+Additionally, the detail list of hyperparamaters that were fixed and tuned can be found in Table \@ref(tab:fixedhyperparam) and \@ref(tab:tunedhyperparam), respectively.
 We applied five-fold cross validation at each combination and recorded model performance on the validation data at each iteration.
-Code for this can be found ^[Ruth, where?].
-Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam)
+Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
+Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam).
 
@@
 With best performing models from hyperparameter scans we carried out a process of fine tuning manually to ensure that models did not over- or underfit data and were trained to be as general as possible.
-We loaded best performing models and assessed changes in selected parameters as described in Results to produce final models ^[Ruth, did you use anything like tensorboard or just R to manually tune the models?].
+We loaded best performing models and assessed changes in selected parameters as described in Results to produce final models.
+These processes were performed manually using Python scripts that can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).
+The results were visualised using R and the codes are stored in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/r-scripts/model-scripts/multi-class/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/r-scripts/model-scripts/multi-class/secreted_data).
 Once each model was tuned accordingly using the training and test sets, final classification testing and estimation of accuracy was done with the held-out validation set that the model had not seen until this point.
-The final models were saved into HDF5 objects and stored in the repository at [link](link) ^[Ruth, link please!]
+The final models were saved into HDF5 objects and stored in the repository at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5).
 
@@
 ```
+
 ```{r, fungiefftable}
@@
 
+```{r, possiblecombination}
+apa_table(
+  readr::read_tsv("tables/possible_combination_each_model.tab"),
+  digits = 0,
+  align = 'c',
+  caption = 'List of possible combination of hyperparamaters setting for each model'
+)
+```
+
+```{r, fixedhyperparam}
+apa_table(
+  readr::read_tsv("tables/fixed_hyperparam.tab"),
+  digits = 0,
+  align = 'c',
+  caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
+)
+```
+
+```{r, tunedhyperparam}
+apa_table(
+  readr::read_tsv("tables/tuned_hyperparam.tab"),
+  digits = 0,
+  align = 'c',
+  caption = 'Hyperparameters tuned in automatic scanning of hyperparameter space for each model'
+)
+```
+
 ```{r, hyperparam }
@@
 ```
+
 # Discussion

### 2020-05-15 main.Rmd
Training and characteristics section

--- 2020-05-11 main.Rmd
@@
 The sequences collected were encoded using either one-hot encoding (CNN-LSTM based models) or integer based encoding (CNN-GRU-LSTM models).
+Sequences were post-padded with zeroes to bring the vectors to identical lengths to each other and the longest sequence in the taxon data set ^[Ruth, what were the lengths for each model?].
 Encoded sequences were split into taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd)
@@
 We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding.
-We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R ^[Ruth, Version?], Tensorflow ^[Ruth, Version?], Keras ^[Ruth, Version?], Python ^[Ruth, Version?] on machines ^[Ruth, Machine types] with  Gb RAM ^[Ruth, Amounts?].
+We trained each model using a basic random hyperparameter setting initialisation step followed by manual tuning.
+All training was performed in R ^[Ruth, Version?], Tensorflow ^[Ruth, Version?], Keras ^[Ruth, Version?], Python ^[Ruth, Version?] on machines ^[Ruth, Machine types] with  Gb RAM ^[Ruth, Amounts?].
 
@@
 Is there a script to reference?].
-We applied five-fold cross validation at each combination and recorded model performance on the validation data at each iteration.
+We applied five-fold cross validation at each combination and recorded model performance on the validation data set at each iteration.
 Code for this can be found ^[Ruth, where?].
-Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam)
 
@@
 
-## Model Training
+## Model Selection and Training
 
-TODO:
+We selected four base architectures on which to build models for learning.
+Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU.
+All models had fully-connected dense layers after this.
+See Figure \@ref(fig:modelarch).
+Hyperparameter values for the models were identified using a random search as described in methods.
+Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
+For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.
+Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), ^[Ruth, what were the actual best accuracy and loss values for the hyperparamter scan models selected and in Table \@ref(tab:hyperparam) , this table should have a column that shows this] a total of twelve models were produced, one for each taxon and model combination.
 
-comes to fig  21, 22, 23
+We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality.
+Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size.
+We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.
+Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in table \@ref(tab:overall_model).
+All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
+
+## Model Characteristics
+
+We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.\@ref(fig:confusionfinalmodels) The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors.
+The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors.
+The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct.
+Overall the models are accurate and show little to no bias with the exception of the GRU-Embedding type.
+In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.
+
+Correlations between the ^[Add correlation to methods] calls \@ref(fig:model_corrs)
+
+## Ensemble Models
+
+## Ensemble Model Characteristics
+
+## Comparison With Other Classification Software
+
+## Convolution Heatmaps
+
+## Software Implementation
 
@@
 ```
+
 ```{r, oomefftable}
@@
 ```
+
+```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
+knitr::include_graphics("figures/model_arch.png")
+```
+
+```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+knitr::include_graphics("figures/final_train_progress.png")
+```
+
+```{r, overallmodel}
+apa_table(
+  readr::read_tsv("tables/final_model_performance.tab"),
+  digits = 3,
+  align = 'c',
+  caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
+)
+```
+
+```{r, confusionfinalmodels, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+
+knitr::include_graphics("figures/final_confusion.png")
+
+```
+
+```{r, modelcorrs, fig.cap="model correlations"}
+knitr::include_graphics("figures/drafts.png")
+```
+
 # Discussion

### 2020-05-21 main.Rmd

@@
 We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding.
-We trained each model using a basic random hyperparameter setting initialisation step followed by manual tuning.
-All training was performed in R ^[Ruth, Version?], Tensorflow ^[Ruth, Version?], Keras ^[Ruth, Version?], Python ^[Ruth, Version?] on machines ^[Ruth, Machine types] with  Gb RAM ^[Ruth, Amounts?].
+We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM.
 
@@
 
-We performed a random search of hyperparameter effects by selecting ^[Ruth, how many?
-The report says 50, but it isn't clear whether this is 50 from 100 possible, or 50 from 50 possible - when you write something you need to keep in mind the question - 'How would my reader repeat this?' ] of all possible hyperparameters for each model type at random ^[Ruth, how is this done?
-Is there a script to reference?].
-We applied five-fold cross validation at each combination and recorded model performance on the validation data set at each iteration.
-Code for this can be found ^[Ruth, where?].
+We performed a random search of hyperparameter effects by selecting 50 of all possible hyperparameters depending on each model type at random.
+The total of possible hyperparameter combination for each model can be found in Table \@ref(tab:possiblecombination).
+Additionally, the detail list of hyperparamaters that were fixed and tuned can be found in Table \@ref(tab:fixedhyperparam) and \@ref(tab:tunedhyperparam), respectively.
+We applied five-fold cross validation at each combination and recorded model performance on the validation data at each iteration.
+Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
+Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam).
 

### 2020-05-21 main.Rmd
add ensemble

@@
 
+# Introduction
+
+Phytopathogens are a major threat to global crop production.
+The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of ~6 billion USD to potato production, annually [@haas2009genome].
+The bacterium _Ralstonia solanacearum_ has a wide host range and can cause loses of over 30% in potato, banana and groundnut [@Yuliar:2015hp].
+The incidences of crop disease are increasing, global climate change and agricultural practice are expanding the geographical range of pathogens and upping the stakes in the evolutionary arms race.
+Effector proteins are the shock troops of infection, manipulating the host at the infection interface to the pathogens advantage.
+Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are **notoriously difficult to characterise from sequence data**.
+In most phyla they have only a few easily determined sequence characteristics (some in fungi are cysteine rich or have a MAX motif, some in oomycetes have the RXLR motif or WY fold)  but in many cases no sequence identifiers are known [@Franceschetti:2017gc].
+Characterising effectors requires painstaking molecular experimental work and genome-scale approaches have relied on complex computational pipelines with in-built _a priori_ assumptions about what might constitute an effector sequence in the absence of sequence features known to group them [@Sperschneider:2015kb].
+To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a critical need to identify effectors computationally from genome and protein sequence data.
+
+Deep learning models can uncover cryptic patterns in data.
+We believe such models can be trained to identify effectors directly from sequence as others have shown it is possible to classify other eukaryote and bacterial nuclear proteins [@Savojardo:2017eq].
+In doing so these models encode information about the features it is using to classify that can be extracted to reveal novel sequence features that current techniques cannot.
+Deep learning models are distinct from other machine learning processes in that pre-selection of important features is far less critical and the models can learn these features unsupervised from training data [@Jurtz:2017cr].
+This property removes the need to know which properties of a data set must be examined before data collection begins.
+
+Convolutional neural networks (CNNs) ...
+
+Recent work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy.
+The work shows clearly that there are unknown patterns in the sequence data that a deep learning method could use to classify.
+
+Here we use combinations of Convolutional Neural Networks (CNNs) that do not rely so heavily on _a priori_ feature selection - only a well annotated data set,  and that can extract novel features of the sequence in an unbiased, more unsupervised manner.
+
+Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as effector or not.
+The larger and more accurate the list the more sensitivity a model can obtain.
+It is critical that training examples are experimentally verified effectors.
+Much of the effector annotation in public genomics databases is from computational predictions of genomics and therefore unverified hypothetical effectors.
+A good source of experimentally verified data is in the Molecular Plant Microbe Interactions (MPMI) literature.
+The widest ranging manual curation of MPMI papers is being performed as part of the PHI-Base [@Urban:2017gp] database strategy.
+
 # Methods
@@
 
-## Procedure
+### Model Classification Correlation
 
-## Data analysis
-We used `r cite_r("r-references.bib")` for all our analyses.
+We calculated correlations between the classifications from best performing models on the hold-out test data set using Pearson's correlation co-efficient on the 1/0 classification vectors.
+
+## Ensemble Functions
+
+We computed an aggregate classification using two different ensemble functions, weighted average and an overall majority option.
+
+Weighted average is computed as
+
+\begin{equation}
+\bar{y} = \frac{\sum_{i=1}^{n} w_i y_i}{\sum_{i=1}^{n} w_i} ,
+\end{equation}
+
+where $w_i$ is the weight, $y_i$ is the prediction value of the $i^{th}$ model, and $n$ is the total number of model. In our case, we use the accuracy of each model as the average.
+
+Overall majority is computed as
+
+\begin{equation}
+\tilde{y} = argmax(x_1, x_2, ..., x_n) ,
+\end{equation}
+where
+
+\begin{equation}
+y =
+   \begin{cases}
+          1 & \text{for } \tilde{y} > 0.5 \\
+          0 & \text{for } \tilde{y} \leq 0.5
+  \end{cases}
+\end{equation}
+
+## Metrics
+
+We used the following calculations for different accuracy metrics in our evaluations, specifically: accuracy, sensitivity, specificity.
+$TP$, $TN$, $FP$, and $FN$ refer to the number of true positives, true negatives, false positives and false negatives, respectively.
+
+Accuracy ($Acc$) is the ratio between correctly classified non-effectors and effectors and all samples:
+\begin{equation}
+  Acc = \frac{TP + TN}{TP + TN + FP + FN}
+\end{equation}
+
+Sensitivity $Sn$ is the ratio between correctly predicted as effectors and all effectors:
+\begin{equation}
+  Sn = \frac{TP}{TP + FN}
+\end{equation}
+
+Specificity ($Sp$) is the ratio between correctly predicted as non-effectors and all non-effectors:
+\begin{equation}
+  Sp = \frac{TN}{TN + FP}
+\end{equation}
 
@@
 
-Correlations between the ^[Add correlation to methods] calls \@ref(fig:model_corrs)
+Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:model_corrs).
+CNN-GRU and LSTM-Embedding showed identical prediction sets.
+We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 - 0.9; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 - 0.51.
+The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.3 - 0.65.
+The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves.
+For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others.
+The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set.
+It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.
 
@@
 
-## Ensemble Model Characteristics
+We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input.
+We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Table \@ref(tab:ensembleresults).
+With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
+However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures.
+The
 
 ## Comparison With Other Classification Software
+Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software
+
+^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document]
 
@@
 
+```{r, ensembleresults}
+apa_table(
+  readr::read_tsv("tables/ensemble_model_performance.tab"),
+  digits = 3,
+  align = 'c',
+  caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
+)
+```
+
 ```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
@@
 
-```{r, modelcorrs, fig.cap="model correlations"}
+```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
 knitr::include_graphics("figures/drafts.png")

### 2020-05-22 main.Rmd
First draft of methods and results

@@
 
+F1-score is the harmonic average of the precision and recall:
+\begin{equation}
+  F_{1} = \frac{2TP}{2TP + FN + FP}
+\end{equation}
+
+## Activation Map Analysis
+
+To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we used a GRAD-CAM approach ^[Ruth, there is absolutely no description of this in your report, please add a description of what you did, with which inputs and settings and versions etc].
+
 # Results
@@
 Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
-We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in table \@ref(tab:overall_model).
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in Fig \@ref(fig:overallmodel).
 All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
@@
 
-Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:model_corrs).
+Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:modelcorrs).
 CNN-GRU and LSTM-Embedding showed identical prediction sets.
@@
 We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input.
-We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Table \@ref(tab:ensembleresults).
+We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults).
 With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
 However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures.
-The
+Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models.
+This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.
 
 ## Comparison With Other Classification Software
-Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software
 
-^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document]
+Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models ^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document] to evaluate the performance of our models against widely used effector identification software.
+We compared against predictions from the bacterial effector prediction programs deepT3 and effectiveT3, the fungal effector prediction programs EffectorP 1 and 2 and the oomycete effector predictor effectR.
+Each comparison was carried out using the respective hold-out test sequence set for each taxon.
+For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics).
+This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
+Absoloute improvements were up to ~ 18 % in bacterial sequences, ~15 % in fungal sequences and ~ 28% in the oomycete sequences.
+The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
+All the established software classifiers we tested show lower sensitivity than the models we have developed here.
 
@@
 
+An advantage of CNNs relative to other deep neural networks is their relative interpretability.
+A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
+To examine the responses of the model, we ran the effector (not the non-effector) ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another].
+data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method , the profiles were smoothed using FFT and examined \@ref(fig:activations).
+All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals.
+The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids.
+Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs even within a taxon and data set.
+The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory ^[Ruth, we need to discuss this bit again, we may need to re-run on individual sets, ie training sets with just effectors in...]
+
 ## Software Implementation
+
+To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the code.
+The package is available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
+
+^[Ruth, will we do a shiny tool?]
 
@@
   align = 'c',
-  caption = 'Effectors found in three major pathogen groups'
+  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
 )
 ```
+
+```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
+knitr::include_graphics("figures/model_arch.png")
+```
+
+```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+knitr::include_graphics("figures/final_train_progress.png")
+```
+
+```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
+# apa_table(
+#   readr::read_tsv("tables/final_model_performance.tab"),
+#   digits = 3,
+#   align = 'c',
+#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
+# )
+readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(Model, Accuracy) +
+  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
+  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
+
+```
+
+```{r, confusionfinalmodels, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+
+knitr::include_graphics("figures/final_confusion.png")
+
+```
+
+```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+knitr::include_graphics("figures/drafts.png")
+```
+
+```{r, ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+# apa_table(
+#   readr::read_tsv("tables/ensemble_model_performance.tab"),
+#   digits = 3,
+#   align = 'c',
+#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
+readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(Combination, Accuracy) +
+  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
+  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
+
+```
+
+```{r, vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
+readr::read_tsv("tables/cfothertools.tab") %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(Classifier, Score) +
+  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
+  ggplot2::facet_wrap(~ Group, scales = "free_x")
+```
+
+```{r, vsotherscfmat, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
+knitr::include_graphics("figures/cf_other_tools.png")
+```
+
+```{r, activations, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines.
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+knitr::include_graphics("figures/convolution_maps.png")
+```
+
+# Discussion
+
+\newpage
+
+# References
+```{r create_r-references}
+r_refs(file = "r-references.bib")
+```
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+<div id="refs" custom-style="Bibliography"></div>
+\endgroup
+
+\newpage
+
+# Supplementary Tables
 
@@
 ```
-
-```{r, ensembleresults}
-apa_table(
-  readr::read_tsv("tables/ensemble_model_performance.tab"),
-  digits = 3,
-  align = 'c',
-  caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
-)
-```
-
-```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
-knitr::include_graphics("figures/model_arch.png")
-```
-
-```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
-knitr::include_graphics("figures/final_train_progress.png")
-```
-
-```{r, overallmodel}
-apa_table(
-  readr::read_tsv("tables/final_model_performance.tab"),
-  digits = 3,
-  align = 'c',
-  caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
-)
-```
-
-```{r, confusionfinalmodels, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
-
-knitr::include_graphics("figures/final_confusion.png")
-
-```
-
-```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
-knitr::include_graphics("figures/drafts.png")
-```
-
-# Discussion
-
-\newpage
-
-# References
-```{r create_r-references}
-r_refs(file = "r-references.bib")
-```
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id="refs" custom-style="Bibliography"></div>
-\endgroup

### 2020-05-31 main.Rmd
Figures are now rendered from data (#1, #2, #3, #4)

@@
 library("papaja")
-library(magrittr)
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+
+# Source plotting functions
+source("functions/plot_confusion_mat.R")
+source("functions/plot_confusion_matrices.R")
+source("functions/get_lower_corr_mat.R")
+source("functions/get_seqlogo.R")
+
+# ggplot2 global theme
+theme_set(theme_bw())
 ```
@@
 
-```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
 knitr::include_graphics(
@@
 
-```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
 knitr::include_graphics(
@@
 
-```{r, effsummary}
+```{r effsummary}
 apa_table(
@@
 
-```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
+```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
 knitr::include_graphics("figures/model_arch.png")
@@
 
-```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+```{r trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
 knitr::include_graphics("figures/final_train_progress.png")
@@
 
-```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
 # apa_table(
@@
   ggplot2::facet_wrap(~ Taxon, scales = "free_x") 
-
 ```
 
-```{r, confusionfinalmodels, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
+```{r confusionfinalmodels, eval=FALSE, include=FALSE, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
 A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
@@
 knitr::include_graphics("figures/final_confusion.png")
-
 ```
 
-```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+```{r modelcorrs, eval=FALSE, include=FALSE, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
 knitr::include_graphics("figures/drafts.png")
@@
 
-```{r, ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models.
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+# Load data
+bacteria_ensemble_results <- data.table::fread("data/raw/bacteria_pred_ens.csv", drop = "V1")
+
+bacteria_test_label <- data.table::fread("data/raw/bacteria_testing.csv") %>%
+  dplyr::select(label) %>%
+  mutate_each(list(as.factor))
+
+oomycete_ensemble_results <- data.table::fread("data/raw/oomycete_pred_ens.csv",drop = "V1")
+
+oomycete_test_label <- data.table::fread("data/raw/oomycete_testing.csv") %>%
+  dplyr::select(label) %>%
+  mutate_each(list(as.factor))
+
+fungi_ensemble_results <- data.table::fread("data/raw/fungi_pred_ens.csv", drop = "V1")
+
+fungi_test_label <- data.table::fread("data/raw/fungi_testing.csv") %>%
+  dplyr::select(label) %>%
+  mutate_each(list(as.factor))
+
+# Correlation plots
+corr_bacteria <- get_lower_corr_mat(bacteria_ensemble_results)
+corr_fungi <- get_lower_corr_mat(fungi_ensemble_results)
+corr_oomycete <- get_lower_corr_mat(oomycete_ensemble_results)
+
+# Confusion matrices
+conf_bacteria <- plot_confusion_matrices(
+  data = bacteria_ensemble_results,
+  true_label = bacteria_test_label,
+  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
+  fill_upper_limit = NA
+)
+
+conf_fungi <- plot_confusion_matrices(
+  data = fungi_ensemble_results,
+  true_label = fungi_test_label,
+  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
+  fill_upper_limit = NA
+)
+
+conf_oomycete <- plot_confusion_matrices(
+  data = oomycete_ensemble_results,
+  true_label = oomycete_test_label,
+  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
+  fill_upper_limit = NA
+)
+
+# Patchwork
+corr_patch <- (
+  corr_bacteria + conf_bacteria +
+    plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
+  ) /
+  (
+    corr_fungi + conf_fungi +
+      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
+  ) /
+  (
+    corr_oomycete + conf_oomycete +
+      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
+  )
+
+corr_patch +
+  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
+```
+
+```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
 # apa_table(
@@
   ggplot2::facet_wrap(~ Taxon, scales = "free_x")
-
 ```
 
-```{r, vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
+```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
 readr::read_tsv("tables/cfothertools.tab") %>% 
@@
 
-```{r, vsotherscfmat, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
+```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
 A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
-knitr::include_graphics("figures/cf_other_tools.png")
+# knitr::include_graphics("figures/cf_other_tools.png")
+
+bacteria_confusion_mat_df <- readr::read_csv("data/processed/bacteria_confusion_mat_df.csv")
+fungi_confusion_mat_df <- readr::read_csv("data/processed/fungi_confusion_mat_df.csv")
+oomycete_confusion_mat_df <- readr::read_csv("data/processed/oomycete_confusion_mat_df.csv")
+
+bacteria_conf <- plot_confusion_mat(
+  bacteria_confusion_mat_df, 
+  model_list = c("ensemble_weighted", "deepT3", "effectiveT3"), 
+  fill_upper_limit = 40
+)
+
+fungi_conf <- plot_confusion_mat(
+  fungi_confusion_mat_df, 
+  model_list = c("cnn_lstm", "effectorp_1", "effectorp_2"),  
+  fill_upper_limit = 40
+)
+
+oomycete_conf <- plot_confusion_mat(
+  oomycete_confusion_mat_df,
+  model_list = c("cnn_lstm", "effectR"),  
+  fill_upper_limit = 40
+)
+
+(bacteria_conf + ggplot2::theme(legend.position = "none") +
+    fungi_conf + 
+    oomycete_conf + ggplot2::theme(legend.position = "none")) +
+   plot_layout(ncol = 1, heights = c(1,1,1)) +
+   plot_annotation(tag_levels = "A")
 ```
 
-```{r, activations, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines.
+```{r activations, fig.height=6, fig.width=10, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines.
 A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
-knitr::include_graphics("figures/convolution_maps.png")
+# knitr::include_graphics("figures/convolution_maps_fixed.pdf")
+
+all_bacteria <- rbind(
+  data.table::fread("data/raw/bacteria_training.csv"),
+  data.table::fread("data/raw/bacteria_validation.csv"),
+  data.table::fread("data/raw/bacteria_testing.csv")
+)
+
+all_oomycete <- rbind(
+  data.table::fread("data/raw/oomycete_training.csv"),
+  data.table::fread("data/raw/oomycete_validation.csv"),
+  data.table::fread("data/raw/oomycete_testing.csv")
+)
+
+all_fungi <- rbind(
+  data.table::fread("data/raw/fungi_training.csv"),
+  data.table::fread("data/raw/fungi_validation.csv"),
+  data.table::fread("data/raw/fungi_testing.csv")
+)
+
+bacteria_heatmap <- data.table::fread("data/raw/bacteria_heatmap.csv")
+fungi_heatmap <- data.table::fread("data/raw/fungi_heatmap.csv")
+oomycete_heatmap <- data.table::fread("data/raw/oomycete_heatmap.csv")
+
+bacteria_eff_seqlogo <- get_seqlogo(
+  all_bacteria %>% dplyr::filter(label == 1), 
+  method_plot = "bits", 
+  start_pos = 25, 
+  end_pos = 50, 
+  break_size = 5
+)
+
+bacteria_non_eff_seqlogo <- get_seqlogo(
+  all_bacteria %>% dplyr::filter(label == 0), 
+  method_plot = "bits", 
+  start_pos = 25, 
+  end_pos = 50, 
+  break_size = 5
+)
+
+oomycete_eff_seqlogo <- get_seqlogo(
+  all_oomycete %>% dplyr::filter(label == 1), 
+  method_plot = "bits", 
+  start_pos = 15, 
+  end_pos = 40, 
+  break_size = 5
+)
+
+oomycete_non_eff_seqlogo <- get_seqlogo(
+  all_oomycete %>% dplyr::filter(label == 0), 
+  method_plot = "bits", 
+  start_pos = 15, 
+  end_pos = 40, 
+  break_size = 5
+)
+
+fungi_eff_seqlogo <- get_seqlogo(
+  all_fungi %>% dplyr::filter(label == 1), 
+  method_plot = "bits", 
+  start_pos = 35, 
+  end_pos = 80, 
+  break_size = 5
+)
+
+fungi_non_eff_seqlogo <- get_seqlogo(
+  all_fungi %>% dplyr::filter(label == 0), 
+  method_plot = "bits", 
+  start_pos = 35, 
+  end_pos = 80, 
+  break_size = 5
+)
+
+heatmap_linebar_bacteria <- bacteria_heatmap %>% 
+  dplyr::mutate(index = V1 + 1) %>% 
+  dplyr::filter(index <= 750) %>% 
+  dplyr::select(index, heatmap_bacteria_smooth) %>% 
+  ggplot(aes(x = index, y = heatmap_bacteria_smooth)) +
+  geom_line() +
+  geom_vline(xintercept = 25, linetype = 'dashed', color = "red") +
+  geom_vline(xintercept = 50, linetype = 'dashed',  color = "red") +
+  scale_x_continuous(breaks = seq(0, 750, 150)) +
+  labs(x = 'Position', y = "Value")
+
+heatmap_linebar_oomycete <- oomycete_heatmap %>% 
+  dplyr::mutate(index = V1 + 1) %>% 
+  dplyr::filter(index <= 750) %>% 
+  dplyr::select(index, heatmap_oomycete_smooth) %>% 
+  ggplot(aes(x = index, y = heatmap_oomycete_smooth)) +
+  geom_line() +
+  geom_vline(xintercept = 15, linetype = 'dashed', color = "blue") +
+  geom_vline(xintercept = 40, linetype = 'dashed', color = "blue") +
+  geom_vline(xintercept = 45, linetype = 'dashed', color = "red") +
+  geom_vline(xintercept = 70, linetype = 'dashed', color = "red") +
+  scale_x_continuous(breaks = seq(0, 750, 150)) +
+  labs(x = 'Position', y = "Value")
+
+heatmap_linebar_fungi <- fungi_heatmap %>%
+  dplyr::mutate(index = V1 + 1) %>% 
+  dplyr::filter(index <= 750) %>% 
+  dplyr::select(index, heatmap_fungi_smooth) %>% 
+  ggplot(aes(x = index, y = heatmap_fungi_smooth)) +
+  geom_line() +
+  geom_vline(xintercept = 35, linetype = 'dashed', color = "blue") +
+  geom_vline(xintercept = 80, linetype = 'dashed', color = "blue") +
+  scale_x_continuous(breaks = seq(0, 750, 150)) +
+  labs(x = 'Position', y = "Value")
+
+seqlogo_patch <- (
+  heatmap_linebar_bacteria + 
+    bacteria_eff_seqlogo + theme(legend.position = "none") +
+    bacteria_non_eff_seqlogo + theme(legend.position = "none") +
+    plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
+  ) /
+  (
+  heatmap_linebar_fungi + 
+    fungi_eff_seqlogo + theme(legend.position = "none") +
+    fungi_non_eff_seqlogo +
+      plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
+  ) /
+  (
+  heatmap_linebar_oomycete + 
+    oomycete_eff_seqlogo + theme(legend.position = "none") +
+    oomycete_non_eff_seqlogo + theme(legend.position = "none") +
+      plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
+  )
+
+seqlogo_patch +
+  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
 ```
@@
 
-```{r, bacefftable}
+```{r bacefftable}
 apa_table(
@@
 
-```{r, fungiefftable}
+```{r fungiefftable}
 apa_table(
@@
 
-```{r, oomefftable}
+```{r oomefftable}
 apa_table(
@@
 
-```{r, possiblecombination}
+```{r possiblecombination}
 apa_table(
@@
 
-```{r, fixedhyperparam}
+```{r fixedhyperparam}
 apa_table(
@@
 
-```{r, tunedhyperparam}
+```{r tunedhyperparam}
 apa_table(
@@
 
-```{r, hyperparam }
+```{r hyperparam}
  readr::read_tsv("tables/hyperparamscan_info.tab") %>% 

### 2020-05-31 data/processed/bacteria_conf_mat_processing.R
Added data processed from data/raw for the manuscript

@@ -1,0 +1,33 @@
+library(tidyverse)
+
+bacteria_test_effectiveT3 <- data.table::fread(
+  "data/raw/bacteria_pred_effectiveT3_testing.csv", drop = "V1"
+) %>% 
+  dplyr::arrange(desc(label)) %>% 
+  dplyr::select(-c("identifier")) %>% 
+  `colnames<-`(c("effectiveT3", "true_label"))
+
+bacteria_test_deept3 <- data.table::fread(
+  "data/raw/bacteria_pred_DeepT3_testing.csv", 
+  drop = "V1"
+) %>% 
+  `colnames<-`("deepT3")
+
+bacteria_ensemble_results <- data.table::fread(
+  "data/raw/bacteria_pred_ens.csv",
+  drop = c("V1", "sequence")
+)
+
+all_bacteria_pred <- bacteria_ensemble_results %>%
+    cbind(., bacteria_test_deept3, bacteria_test_effectiveT3)
+
+model_list <- c("ensemble_weighted", "deepT3", "effectiveT3")
+
+confusion_mat_bacteria_df <- all_bacteria_pred %>% 
+  dplyr::mutate_all(function(x) factor(x, levels = c(1, 0))) %>% 
+  tidyr::pivot_longer(-true_label, names_to = "model", values_to = "Prediction") %>% 
+  dplyr::mutate(model = factor(model, levels = model_list)) %>% 
+  table() %>%
+  as.data.frame()
+
+readr::write_csv(confusion_mat_bacteria_df, "data/processed/bacteria_confusion_mat_df.csv")

### 2020-05-31 data/processed/oomycete_conf_mat_processing.R

@@ -1,0 +1,16 @@
+library(tidyverse)
+
+all_oomycete_pred <-  data.table::fread("data/raw/oomycete_pred_all.csv", drop = c("V1", "sequence"))
+
+model_list <- c("cnn_lstm", "effectR")
+
+# Get the confusion matrix data
+confusion_mat_oomycete_df <- all_oomycete_pred %>% 
+  dplyr::mutate_all(function(x) factor(x, levels = c(1, 0))) %>% 
+  dplyr::select(-c(ensemble_weighted)) %>% 
+  tidyr::pivot_longer(-true_label, names_to = "model", values_to = "Prediction") %>% 
+  dplyr::mutate(model = factor(model, levels = model_list)) %>% 
+  table() %>%
+  as.data.frame()
+
+readr::write_csv(confusion_mat_oomycete_df, "data/processed/oomycete_confusion_mat_df.csv")

### 2020-05-31 functions/plot_confusion_matrices.R
Added functions to produce plots

@@ -1,0 +1,32 @@
+plot_confusion_matrices <- function(data, true_label, model_list, nrow = 1, fill_upper_limit = NA) {
+  conf_matrix_df <- data %>%
+    select(-sequence) %>%
+    # Add true labels
+    dplyr::mutate(Reference = true_label$label) %>%
+    # Transform into factors
+    dplyr::mutate_all(function(x) factor(x, levels = c(1, 0))) %>%
+    # Select chosen model only
+    tidyr::pivot_longer(-Reference, names_to = "model", values_to = "Prediction") %>%
+    # Filter models
+    dplyr::filter(model %in% model_list) %>% 
+    dplyr::mutate(model = factor(model, levels = model_list)) %>%
+    # Calculate frequencies
+    table() %>%
+    as.data.frame()
+  
+  # Make plot
+  gg_matrix <- conf_matrix_df %>%
+    ggplot() +
+    aes(x = Reference, y = Prediction) +
+    geom_tile(aes(fill = Freq), colour = "white") +
+    geom_text(aes(label = Freq), vjust = 0.5) +
+    # scale_fill_gradient(low = "lightpink", high = "mediumpurple1") +
+    scale_fill_viridis_c(begin = 0.1, end = 1, limits = c(0, fill_upper_limit)) +
+    labs(x = "True value", y = "Prediction") +
+    coord_fixed() +
+    facet_wrap(~model, nrow = nrow) +
+    theme_bw() +
+    theme(legend.position = "none")
+  
+  return(gg_matrix)
+}

### 2020-06-04 main.Rmd
edits

--- 2020-05-22 main.Rmd
@@
 ---
-title             : "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
-shorttitle        : "CNNs for classifying effector proteins"
-
+title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
+date: "`r format(Sys.time(), '%d %B, %Y')`"
+output: 
+  bookdown::pdf_document2:
+    citation_package: natbib
+    keep_tex: true
+    fig_caption: true
+    latex_engine: pdflatex
+    template: svm_latex.tex
+    toc: no
+fontsize: 12pt
+fontfamily: sans
+papersize: a4paper
 author:
-  - name          : "Ruth Kristianingsih"
-    affiliation   : "1"
-    email         : "ruth.kristianingsih@tsl.ac.uk"
-  - name          : "Dan MacLean"
-    email         : "dan.maclean@tsl.ac.uk"
-    address       : "The Sainsbury Laboratory, University of East Anglia, Norwich, UK, NR4 7UH"
-    corresponding : yes    # Define only one corresponding author
-    affiliation   : "1"
-
-affiliation:
-  - id            : "1"
-    institution   : "The Sainsbury Laboratory"
-
-abstract: |
-  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
-
-  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
-
-  One sentence clearly stating the **general problem** being addressed by  this particular study.
-
-  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
-
-  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
-
-  One or two sentences to put the results into a more **general context**.
-
-  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
-
-  <!-- https://tinyurl.com/ybremelq -->
-
-  <!-- keywords          : "keywords" 
-  wordcount         : "X" -->
-
-bibliography      : ["r-references.bib"]
-
-floatsintext      : no
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-linenumbers       : yes
-mask              : no
-draft             : no
-
-documentclass     : "apa6"
-classoption       : "man"
-output            : papaja::apa6_pdf
+  - name: Ruth Kristianingsih, Dan MacLean
+  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
+abstract: "TBD."
+geometry: margin=1in
+biblio-style: apsr
+bibliography: "r-references.bib"
+graphics: yes
+header-includes:
+  - \usepackage{amsmath}
 ---
@@
 
-```{r analysis-preferences}
+```{r analysis-preferences, echo=FALSE}
 # Seed for random number generation
@@
 We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn.
-We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates_ensembl_2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley_biomart_2009].
+We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley2009].
 Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP.
-With these we used SignalP 3.0 [@dyrlov_bendtsen_improved_2004] in order to filter the secreted sequences and selected accordingly.
-^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camacho_blast_2009].
+With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly.
+^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camachoblast2009].
 We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
 
-```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt", echo=FALSE}
 knitr::include_graphics(
@@
 
-```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists", echo=FALSE}
 knitr::include_graphics(
@@
    \begin{cases}
-          1 & \text{for } \tilde{y} > 0.5 \\
-          0 & \text{for } \tilde{y} \leq 0.5
+          1 & \mbox{for } \tilde{y} > 0.5 \\
+          0 & \mbox{for } \tilde{y} \leq 0.5
   \end{cases}
@@
 
-```{r, effsummary}
+```{r, effsummary, echo=FALSE,message = FALSE}
 apa_table(
@@
 
-```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
+```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model", echo=FALSE}
 knitr::include_graphics("figures/model_arch.png")
@@
 
-```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set", echo=FALSE}
 knitr::include_graphics("figures/final_train_progress.png")
@@
 
-```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.", echo=FALSE, message=FALSE}
 # apa_table(
@@
 ```{r, confusionfinalmodels, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", echo=FALSE}
 
@@
 
-```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", echo=FALSE}
 knitr::include_graphics("figures/drafts.png")
@@
 
-```{r, ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r, ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.", message = FALSE, echo=FALSE}
 # apa_table(
@@
 
-```{r, vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
+```{r, vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", message = FALSE, echo=FALSE}
 readr::read_tsv("tables/cfothertools.tab") %>% 
@@
 ```{r, vsotherscfmat, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
-A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
 knitr::include_graphics("figures/cf_other_tools.png")
@@
 ```{r, activations, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", echo=FALSE}
 knitr::include_graphics("figures/convolution_maps.png")
@@
 
-# References
-```{r create_r-references}
-r_refs(file = "r-references.bib")
-```
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id="refs" custom-style="Bibliography"></div>
-\endgroup
-
-\newpage
-
 # Supplementary Tables
 
-```{r, bacefftable}
+```{r, bacefftable, echo=FALSE, message = FALSE}
 apa_table(
@@
 
-```{r, fungiefftable}
+```{r, fungiefftable, echo=FALSE, message = FALSE}
 apa_table(
@@
 
-```{r, oomefftable}
+```{r, oomefftable, echo=FALSE,message = FALSE}
 apa_table(
@@
 
-```{r, possiblecombination}
+```{r, possiblecombination, echo=FALSE,message = FALSE}
 apa_table(
@@
 
-```{r, fixedhyperparam}
+```{r, fixedhyperparam, echo=FALSE,message = FALSE}
 apa_table(
@@
 
-```{r, tunedhyperparam}
+```{r, tunedhyperparam, echo=FALSE,message = FALSE}
 apa_table(
@@
 
-```{r, hyperparam }
+```{r, hyperparam, echo=FALSE,message = FALSE }
  readr::read_tsv("tables/hyperparamscan_info.tab") %>% 

### 2020-06-04 main.Rmd

--- 2020-05-31 main.Rmd
@@
 We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn.
-We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates_ensembl_2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley_biomart_2009].
+We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley2009].
 Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP.
-With these we used SignalP 3.0 [@dyrlov_bendtsen_improved_2004] in order to filter the secreted sequences and selected accordingly.
-^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camacho_blast_2009].
+With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly.
+^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camachoblast2009].
 We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
 
-```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt", message = FALSE, echo=FALSE}
+
 knitr::include_graphics(
@@
 
-```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists", message = FALSE, echo=FALSE}
+
 knitr::include_graphics(
@@
 
-^[Ruth, will we do a shiny tool?]
+```{r, effsummary, echo=FALSE,message = FALSE}
 
-```{r effsummary}
 apa_table(
@@
 
-```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
+```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model", message = FALSE, echo=FALSE}
 knitr::include_graphics("figures/model_arch.png")
@@
 
-```{r trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set", message = FALSE, echo=FALSE}
 knitr::include_graphics("figures/final_train_progress.png")
@@
 
-```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.", echo=FALSE, message=FALSE}
+
 # apa_table(
@@
 
-```{r confusionfinalmodels, eval=FALSE, include=FALSE, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
-
-knitr::include_graphics("figures/final_confusion.png")
-```
-
-```{r modelcorrs, eval=FALSE, include=FALSE, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
-knitr::include_graphics("figures/drafts.png")
-```
-
 ```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
 # Load data
@@
 
-```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.", message = FALSE, echo=FALSE}
+
 # apa_table(
@@
 
-```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
+```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", message = FALSE, echo=FALSE}
 readr::read_tsv("tables/cfothertools.tab") %>% 
@@
 
-# References
-```{r create_r-references}
-r_refs(file = "r-references.bib")
-```
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id="refs" custom-style="Bibliography"></div>
-\endgroup
-
-\newpage
-
 # Supplementary Tables
 
-```{r bacefftable}
+```{r, bacefftable, echo=FALSE, message = FALSE}
+
 apa_table(
@@
 ```{r fungiefftable}
+
 apa_table(
@@
 
-```{r oomefftable}
+```{r, oomefftable, echo=FALSE,message = FALSE}
+
 apa_table(
@@
 
-```{r possiblecombination}
+```{r, possiblecombination, echo=FALSE,message = FALSE}
+
 apa_table(
@@
 
-```{r fixedhyperparam}
+```{r, fixedhyperparam, echo=FALSE,message = FALSE}
+
 apa_table(
@@
 
-```{r tunedhyperparam}
+```{r, tunedhyperparam, echo=FALSE,message = FALSE}
+
 apa_table(
@@
 
-```{r hyperparam}
+```{r, hyperparam, echo=FALSE,message = FALSE }
+
  readr::read_tsv("tables/hyperparamscan_info.tab") %>% 

### 2020-06-05 main.Rmd
finished intro:

@@
 Effector proteins are the shock troops of infection, manipulating the host at the infection interface to the pathogens advantage.
-Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are **notoriously difficult to characterise from sequence data**.
+Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are notoriously difficult to characterise from sequence data.
 In most phyla they have only a few easily determined sequence characteristics (some in fungi are cysteine rich or have a MAX motif, some in oomycetes have the RXLR motif or WY fold)  but in many cases no sequence identifiers are known [@Franceschetti:2017gc].
 Characterising effectors requires painstaking molecular experimental work and genome-scale approaches have relied on complex computational pipelines with in-built _a priori_ assumptions about what might constitute an effector sequence in the absence of sequence features known to group them [@Sperschneider:2015kb].
-To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a critical need to identify effectors computationally from genome and protein sequence data.
+To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a need to identify effectors computationally from genome and protein sequence data.
 
-Deep learning models can uncover cryptic patterns in data.
-We believe such models can be trained to identify effectors directly from sequence as others have shown it is possible to classify other eukaryote and bacterial nuclear proteins [@Savojardo:2017eq].
-In doing so these models encode information about the features it is using to classify that can be extracted to reveal novel sequence features that current techniques cannot.
+Machine learning (ML) algorithms are a general group of techniques most often used for classification of data into groups.
+Supervised ML require a set of training examples and associated data with which to learn.
+Defining the best data to use and collect, called feature selection is an important and difficult prerequisite.
+ML approaches have been applied with success to biological sequence analysis, particularly in transcription factor binding site prediction, for the classification of eukaryote and bacterial nuclear proteins [@Savojardo:2017eq] and in the plant pathogen domain work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy.
+
 Deep learning models are distinct from other machine learning processes in that pre-selection of important features is far less critical and the models can learn these features unsupervised from training data [@Jurtz:2017cr].
 This property removes the need to know which properties of a data set must be examined before data collection begins.
+The Deep learning models can therefore classify on properties not necessarily known to the operator and could be used to uncover cryptic patterns in data.
+Convolutional neural networks (CNNs) are a type of neural network that have found wide application in numerous machine vision problems, including image object classification and facial identification [@krizhevsky2012, @lawrence1997], in time-series data analysis [@pyrkov2018] and natural language processing [@collobert2008].
+In the biomedical domain they have been used in drug discovery [@wallach2015] and gene network prediction [@maclean2019].
+CNNs encode information about the features used to classify that can be extracted and interpreted.
+In a sequence classification problems this means they have the potential to reveal novel sequence features that other bioinformatics approaches have not and could be of particular utility when analysing sets of effectors.
 
-Convolutional neural networks (CNNs) ...
-
-Recent work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy.
-The work shows clearly that there are unknown patterns in the sequence data that a deep learning method could use to classify.
-
-Here we use combinations of Convolutional Neural Networks (CNNs) that do not rely so heavily on _a priori_ feature selection - only a well annotated data set,  and that can extract novel features of the sequence in an unbiased, more unsupervised manner.
-
-Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as effector or not.
+Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as an effector or not.
 The larger and more accurate the list the more sensitivity a model can obtain.
 It is critical that training examples are experimentally verified effectors.
-Much of the effector annotation in public genomics databases is from computational predictions of genomics and therefore unverified hypothetical effectors.
-A good source of experimentally verified data is in the Molecular Plant Microbe Interactions (MPMI) literature.
-The widest ranging manual curation of MPMI papers is being performed as part of the PHI-Base [@Urban:2017gp] database strategy.
+Much of the effector annotation in public genomics databases is from computational predictions of genomics and is therefore of experimentally unverified hypothetical effectors.
+A good source of experimentally verified data is in the Molecular Plant Microbe Interactions (MPMI) literature and The widest ranging manual curation of MPMI papers is being performed as part of the PHI-Base [@phibase] database strategy, PHI-base is an expertly curated database of genes proven experimentally to affect the outcome of pathogen host interactions and is therefore an excellent source of reliable effector sequences.
+
+Here we use combinations of CNNs that do not rely on _a priori_ feature selection to classify experimentally verified effectors and non-effectors in three taxa of plant pathogen: bacterial, fungal and oomycete.
+We show that these have very strong predictive power, can outperform existing effector prediction methods in accuracy and a better balance of sensitivity and specificity.
+We also analyse the activations of the models in response to effectors and non-effectors to gain insights into the sequence features that are allowing classification.
+We have produced an R package that will allow other scientists to easily classify their own sequences of interest.
 
@@
 The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct.
-Overall the models are accurate and show little to no bias with the exception of the GRU-Embedding type.
+Overall the models are accurate and show little to no bias toward false positive or false negatives, with the exception of the GRU-Embedding type.
 In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.
@@
 ```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
-A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
 # knitr::include_graphics("figures/cf_other_tools.png")
@@
 ```{r activations, fig.height=6, fig.width=10, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
 # knitr::include_graphics("figures/convolution_maps_fixed.pdf")
@@
 
-```{r fungiefftable}
+```{r fungiefftable, message = FALSE, echo=FALSE}
 

### 2020-06-05 main.Rmd
references

@@
 Defining the best data to use and collect, called feature selection is an important and difficult prerequisite.
-ML approaches have been applied with success to biological sequence analysis, particularly in transcription factor binding site prediction, for the classification of eukaryote and bacterial nuclear proteins [@Savojardo:2017eq] and in the plant pathogen domain work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy.
+ML approaches have been applied with success to biological sequence analysis, particularly in transcription factor binding site prediction, for the classification of eukaryote and bacterial nuclear proteins [@Savojardo:2017eq] and in the plant pathogen domain work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy [@effectorp, @effectorp2].
 
@@
 In the biomedical domain they have been used in drug discovery [@wallach2015] and gene network prediction [@maclean2019].
+In studies with bacterial type III secreted effectors Xue et al developed an accurate CNN classifier for bacterial sequences [@deept3].
 CNNs encode information about the features used to classify that can be extracted and interpreted.
@@
 Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models ^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document] to evaluate the performance of our models against widely used effector identification software.
-We compared against predictions from the bacterial effector prediction programs deepT3 and effectiveT3, the fungal effector prediction programs EffectorP 1 and 2 and the oomycete effector predictor effectR.
+We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR].
 Each comparison was carried out using the respective hold-out test sequence set for each taxon.
@@
 
+^[Ruth, these confusion matrices have different axis orders to the ones earlier. Can you make them consistent?]
+
 ## Convolution Heatmaps
@@
 A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
-To examine the responses of the model, we ran the effector (not the non-effector) ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another].
-data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method , the profiles were smoothed using FFT and examined \@ref(fig:activations).
+To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
+The profiles were smoothed using FFT and examined \@ref(fig:activations).
 All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals.
@@
 Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs even within a taxon and data set.
-The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory ^[Ruth, we need to discuss this bit again, we may need to re-run on individual sets, ie training sets with just effectors in...]
+The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory
 
@@
 
-To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the code.
-The package is available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
+To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
+The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
 

### 2020-06-07 data/processed/oomycete_conf_mat_processing.R
Updated scripts used for processing confusion matrices vs other tools

@@ -1,6 +1,31 @@
 library(tidyverse)
 
-all_oomycete_pred <-  data.table::fread("data/raw/oomycete_pred_all.csv", drop = c("V1", "sequence"))
+
+# Read all results from effectR
+pred_result_fungi_effectR <- data.table::fread("data/raw/oomycete_pred_effectR_testing.csv", drop = c("V1")) %>% 
+  dplyr::rename(effectR = pred)
+
+# Get the prediction results from deepredeff
+oomycete_pred_ens <-  data.table::fread("data/raw/oomycete_pred_ens.csv", drop = c("V1"))
+
+# Get the real data 
+oomycete_test_data <- data.table::fread("data/raw/oomycete_testing.csv")
+
+# Combine real data and deepredeff results to get the true label
+
+oomycete_pred_ens <- oomycete_pred_ens %>% 
+  dplyr::left_join(., oomycete_test_data, by = c("sequence" = "Sequence")) 
+
+
+all_oomycete_pred <- oomycete_pred_ens %>% 
+  dplyr::left_join(., pred_result_fungi_effectR, by = c("sequence" = "id", "label" = "true_label")) %>% 
+  # if they are not predicted as effector, then non-effector
+  dplyr::mutate(effectR = ifelse(is.na(effectR), 0, effectR)) %>% 
+  dplyr::rename(true_label = label)
+
+# Save to the model
+readr::write_csv(all_oomycete_pred, "data/processed/all_oomycete_pred_deepredeff_effectR.csv")
+
 
 model_list <- c("cnn_lstm", "effectR")
 

### 2020-06-07 main.Rmd
added Grad-CAM part and remade some plots

@@
     latex_engine: pdflatex
-    template: svm_latex.tex
+    # template: svm_latex.tex
     toc: no
@@
 biblio-style: apsr
-bibliography: "r-references.bib"
+bibliography: ["references.bib", "r-packages.bib"]
 graphics: yes
@@
   - \usepackage{amsmath}
+  - \usepackage{siunitx}
 ---
@@
 source("functions/plot_confusion_matrices.R")
+source("functions/plot_acc_loss_history.R")
 source("functions/get_lower_corr_mat.R")
 source("functions/get_seqlogo.R")
+
+# Build R pacakges Bib build
+r_refs(file = "r-packages.bib")
 
@@
 set.seed(42)
-knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+knitr::opts_chunk$set(
+  cache.extra = knitr::rand_seed,
+  message = FALSE,
+  echo = FALSE
+)
 ```
@@
 We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn.
-We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley2009].
+We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedleybiomart2009].
 Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP.
 With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly.
-^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camachoblast2009].
+We used default paramaters from SignalP, except the type of organism group which is `euk` for both fungi and oomycete sequences, and `gram-` or `gram+` for bacteria sequences.
+Redundant sequences were filtered using BLASTp [@camachoblast2009].
 We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
 
-```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt", message = FALSE, echo=FALSE}
-
+```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
 knitr::include_graphics(
-  "figures/flowchart_effectorsequence_collection.pdf",
+  "figures/flowchart_effectorsequence_collection.pdf"
 )
@@
 
-```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists", message = FALSE, echo=FALSE}
-
+```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
 knitr::include_graphics(
-  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
+  "figures/flowchart_secreted_noneffectorsequence_collection.pdf"
 )
@@
 The sequences collected were encoded using either one-hot encoding (CNN-LSTM based models) or integer based encoding (CNN-GRU-LSTM models).
-Sequences were post-padded with zeroes to bring the vectors to identical lengths to each other and the longest sequence in the taxon data set ^[Ruth, what were the lengths for each model?].
+Sequences were post-padded with zeroes to bring the vectors to identical lengths to each other and the longest sequence in the taxon data set.
+The longest sequence  for bacteria, fungi, and oomycete are 2574, 4034, and 934, respectively.
 Encoded sequences were split into taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd)
@@
 
-To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we used a GRAD-CAM approach ^[Ruth, there is absolutely no description of this in your report, please add a description of what you did, with which inputs and settings and versions etc].
+To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we adapted a Grad-CAM (Gradient-weighted Class Activation Mapping) approach [@Selvaraju2019].
+Using this method, we could produce the visualization of one of the convolutional neural network output values when it does a prediction.
+This will allow us to see which position in peptide sequences is important in determining whether it is effector or non-effector.
+
+To find the importance of sequence position detected by our model, we took the output feature map of a convolution layer, given a sequence input.
+The convolutional layer we chose is the `convd_1` layer.
+Then, we computed the gradient of model output with the respect of the output of convolutional layer we chose.
+After that, we weighed every channel of that feature map with the gradient we calculated earlier, and took an average of the result.
+
+Following the process mentioned above, we then summed all of the activation map results of all sequences in our datasets for each taxon and normalised them.
+Next, the final summed activation map for each taxon was smoothed using fourier analysis and plotted the smoothed version for each of them.
+The code on computing and visualising these heatmaps can be found on [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .
 
@@
 For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.
-Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), ^[Ruth, what were the actual best accuracy and loss values for the hyperparamter scan models selected and in Table \@ref(tab:hyperparam) , this table should have a column that shows this] a total of twelve models were produced, one for each taxon and model combination.
+Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), a total of twelve models were produced, one for each taxon and model combination.
+The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).
 
@@
 Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
-We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in Fig \@ref(fig:overallmodel).
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range $\sim 93.4 \%$  to  $\sim 97.4 \%$, for the four fungal models we observed accuracy in the range $\sim 60.5 \%$  to  $\sim 84.2 \%$  and for the four oomycete models we saw accuracy from $\sim 64.7 \%$  to  $\sim 82.3 \%$, reported in Fig \@ref(fig:overallmodel).
 All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
@@
 
-We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.\@ref(fig:confusionfinalmodels) The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors.
+We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.
+The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors.
 The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors.
@@
 
-Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:modelcorrs).
+Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range \numrange[range-phrase = --]{0.80}{0.88}, see Figure \@ref(fig:confusionmodelcorrs).
 CNN-GRU and LSTM-Embedding showed identical prediction sets.
-We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 - 0.9; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 - 0.51.
-The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.3 - 0.65.
+We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range \numrange[range-phrase = --]{0.79}{0.90}; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range \numrange[range-phrase = --]{0.36}{0.51}.
+The models trained on oomycete sequences all showed this lower range of correlations, in the range \numrange[range-phrase = --]{0.30}{0.65}.
 The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves.
@@
 
-Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models ^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document] to evaluate the performance of our models against widely used effector identification software.
+Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software.
 We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR].
@@
 This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
-Absoloute improvements were up to ~ 18 % in bacterial sequences, ~15 % in fungal sequences and ~ 28% in the oomycete sequences.
+Absoloute improvements were up to $\sim 15\%$  in bacterial sequences, $\sim 15\%$ in fungal sequences and $\sim 20\%$  in the oomycete sequences.
 The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
 All the established software classifiers we tested show lower sensitivity than the models we have developed here.
-
-^[Ruth, these confusion matrices have different axis orders to the ones earlier. Can you make them consistent?]
 
@@
 A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
-To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
+To examine the responses of the model, we ran the effector sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the Grad-CAM method.
 The profiles were smoothed using FFT and examined \@ref(fig:activations).
@@
 To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
-The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
+The package ans installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff).
 
-```{r, effsummary, echo=FALSE,message = FALSE}
-
+```{r, effsummary}
 apa_table(
@@
 
-```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model", message = FALSE, echo=FALSE}
-knitr::include_graphics("figures/model_arch.png")
+```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model"}
+knitr::include_graphics("figures/model_arch.pdf")
 ```
 
-```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set", message = FALSE, echo=FALSE}
-knitr::include_graphics("figures/final_train_progress.png")
+```{r trainingprogress, fig.height= 10, fig.width= 10, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+# knitr::include_graphics("figures/final_train_progress.png")
+
+bacteria_acc_loss <- plot_acc_loss_history(
+  train_data = data.table::fread("data/processed/bacteria_hist_train.csv"),
+  val_data = data.table::fread("data/processed/bacteria_hist_val.csv"),
+  plot_tag = "A"
+) + ggplot2::theme(legend.position = "none", axis.title.x=element_blank())
+
+fungi_acc_loss <- plot_acc_loss_history(
+  train_data = data.table::fread("data/processed/fungi_hist_train.csv"),
+  val_data = data.table::fread("data/processed/fungi_hist_val.csv"),
+  plot_tag = "B"
+) +
+  ggplot2::theme(strip.background.x = element_blank(),
+   strip.text.x = element_blank(),
+    axis.title.x = element_blank())
+
+oomycete_acc_loss <- plot_acc_loss_history(
+  train_data = data.table::fread("data/processed/oomycete_hist_train.csv"),
+  val_data = data.table::fread("data/processed/oomycete_hist_val.csv"),
+  plot_tag = "C"
+) +
+  ggplot2::theme(legend.position = "none") +
+  ggplot2::theme(strip.background.x = element_blank(),
+   strip.text.x = element_blank())
+
+(bacteria_acc_loss + fungi_acc_loss + oomycete_acc_loss) +
+  plot_layout(nrow = 3, heights = c(3, 3, 3))
 ```
 
-```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.", echo=FALSE, message=FALSE}
-
+```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
 # apa_table(
@@
   ggplot2::ggplot() +
-  ggplot2::aes(Model, Accuracy) +
-  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
-  ggplot2::facet_wrap(~ Taxon, scales = "free_x") 
+  ggplot2::aes(
+    x = Model,
+    y = Accuracy,
+    color = Accuracy
+  ) +
+  ggplot2::geom_point() +
+  ggplot2::scale_color_viridis_c(begin = 0.65, end = 0.1) +
+  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
+  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
@@
 ```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
 # Load data
@@
   dplyr::select(label) %>%
-  mutate_each(list(as.factor))
+  dplyr::mutate(label = as.factor(label))
 
@@
   dplyr::select(label)  %>%
-  mutate_each(list(as.factor))
+  dplyr::mutate(label = as.factor(label))
 
@@
   fill_upper_limit = NA
-)
+) +
+  ggplot2::labs(tag = "A")
 
@@
   fill_upper_limit = NA
-)
+) +
+  ggplot2::labs(tag = "B")
 
@@
   fill_upper_limit = NA
-)
+) +
+  ggplot2::labs(tag = "C")
 
@@
 corr_patch <- (
-  corr_bacteria + conf_bacteria +
-    plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
+  conf_bacteria + corr_bacteria +
+    plot_layout(ncol = 2, widths = c(3,1), tag_level = "new")
   ) /
   (
-    corr_fungi + conf_fungi +
-      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
+    conf_fungi +  corr_fungi +
+      plot_layout(ncol = 2, widths = c(3,1), tag_level = "new")
   ) /
   (
-    corr_oomycete + conf_oomycete +
-      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
+    conf_oomycete + corr_oomycete +
+      plot_layout(ncol = 2, widths = c(3,1), tag_level = "new")
   )
@@
 corr_patch +
-  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
+  plot_annotation(tag_levels = c("A"))
 ```
 
-```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.", message = FALSE, echo=FALSE}
-
+```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
 # apa_table(
@@
   ggplot2::ggplot() +
-  ggplot2::aes(Combination, Accuracy) +
-  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
-  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
+  ggplot2::aes(x = Combination, y = Accuracy) +
+  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction,
+                                   shape = EnsembleFunction), size = 2) +
+  ggplot2::scale_shape_manual(values = c("Ensemble Voting" = 19, "Ensemble Weighted" = 17)) +
+  ggplot2::scale_color_viridis_d(begin = 0.65, end = 0.1) +
+  ggplot2::labs(color = "Ensemble Function", shape = "Ensemble Function") +
+  ggplot2::facet_wrap(~ Taxon, scales = "free_x") +
+  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
 ```
 
-```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", message = FALSE, echo=FALSE}
-readr::read_tsv("tables/cfothertools.tab") %>% 
+```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
+data.table::fread("data/processed/acc_f1score_vs_othertools.csv") %>%
+  dplyr::mutate(model = case_when(
+    model == "ensemble_weighted" ~ "Ensemble Weighted",
+    model == "cnn_lstm" ~ "CNN-LSTM",
+    model == "deepT3" ~ "DeepT3",
+    model == "effectiveT3" ~ "EffectiveT3",
+    model == "effectorp_1" ~ "EffectorP 1",
+    model == "effectorp_2" ~ "EffectorP 2",
+    TRUE ~ "EffectR"
+  )) %>%
+  dplyr::mutate(
+    model = factor(model, levels = c("Ensemble Weighted", "DeepT3", "EffectiveT3",
+                                    "CNN-LSTM", "EffectorP 1", "EffectorP 2",
+                                    "EffectR"))
+  ) %>%
   ggplot2::ggplot() +
-  ggplot2::aes(Classifier, Score) +
-  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
-  ggplot2::facet_wrap(~ Group, scales = "free_x")
+  ggplot2::aes(
+    x = model,
+    y = .estimate,
+    color = .metric,
+    shape = .metric
+  ) +
+  ggplot2::scale_shape_manual(values = c("Accuracy" = 19, "F1 Score" = 17)) +
+  ggplot2::geom_point(size = 2) +
+  ggplot2::facet_wrap(~groups, nrow = 1, scales = "free_x") +
+  ggplot2::scale_color_viridis_d(begin = 0.65, end = 0.1) +
+  ggplot2::labs(x = "Classifier", y = "Score", color = "Metric", shape = "Metric") +
+  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
 ```
@@
 ```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
-A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
 # knitr::include_graphics("figures/cf_other_tools.png")
@@
   fill_upper_limit = 40
-)
+) + ggplot2::theme(legend.position = "none")
 
@@
   fill_upper_limit = 40
-)
+) + ggplot2::theme(legend.position = "none")
 
-(bacteria_conf + ggplot2::theme(legend.position = "none") +
-    fungi_conf + 
-    oomycete_conf + ggplot2::theme(legend.position = "none")) +
+(bacteria_conf + fungi_conf + oomycete_conf) +
   plot_layout(ncol = 1, heights = c(1,1,1)) +
@@
 
-```{r activations, fig.height=6, fig.width=10, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
+```{r activations, fig.height=6, fig.width=10, fig.cap="Grad-CAM activation profiles of CNN-LSTM models trained on when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data and non-effector data.
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
 # knitr::include_graphics("figures/convolution_maps_fixed.pdf")
@@
   scale_x_continuous(breaks = seq(0, 750, 150)) +
-  labs(x = 'Position', y = "Value")
+  labs(x = 'Position', y = "Value") +
+  ggplot2::labs(tag = "A")
 
@@
   scale_x_continuous(breaks = seq(0, 750, 150)) +
-  labs(x = 'Position', y = "Value")
+  labs(x = 'Position', y = "Value") +
+  ggplot2::labs(tag = "B")
 
@@
   scale_x_continuous(breaks = seq(0, 750, 150)) +
-  labs(x = 'Position', y = "Value")
+  labs(x = 'Position', y = "Value") +
+  ggplot2::labs(tag = "C")
 
@@
     bacteria_non_eff_seqlogo + theme(legend.position = "none") +
-    plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
+    plot_layout(nrow = 1,  tag_level = "new")
   ) /
@@
     fungi_non_eff_seqlogo +
-      plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
+      plot_layout(nrow = 1,  tag_level = "new")
   ) /
@@
     oomycete_non_eff_seqlogo + theme(legend.position = "none") +
-      plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
+      plot_layout(nrow = 1,  tag_level = "new")
   )
 
-seqlogo_patch +
-  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
+seqlogo_patch
 ```
@@
 
-```{r, bacefftable, echo=FALSE, message = FALSE}
-
+```{r, bacefftable}
 apa_table(
@@
 
-```{r fungiefftable, message = FALSE, echo=FALSE}
-
+```{r, fungiefftable}
 apa_table(
@@
 
-```{r, oomefftable, echo=FALSE,message = FALSE}
-
+```{r, oomefftable}
 apa_table(
@@
 
-```{r, possiblecombination, echo=FALSE,message = FALSE}
-
+```{r, possiblecombination}
 apa_table(
@@
 
-```{r, fixedhyperparam, echo=FALSE,message = FALSE}
-
+```{r, fixedhyperparam}
 apa_table(
@@
 
-```{r, tunedhyperparam, echo=FALSE,message = FALSE}
-
+```{r, tunedhyperparam}
 apa_table(
@@
 
-```{r, hyperparam, echo=FALSE,message = FALSE }
-
+```{r hyperparam}
  readr::read_tsv("tables/hyperparamscan_info.tab") %>%
@@
 ```
+
+```{r acclossfinal}
+ readr::read_tsv("tables/acc_loss_final.tab") %>%
+  knitr::kable( "latex", longtable = T, booktabs = T,
+           digits = 4,
+  align = 'c',
+  caption = 'Accuracy and loss values from best performing parameters values for each model'
+         ) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
+```

### 2020-06-08 main.Rmd
complete hypergeo analysis of regions

--- 2020-06-05 main.Rmd
@@
 The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids.
-Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs even within a taxon and data set.
-The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory
+We examined further the sequences under the largest peaks, specifically for bacterial sequences we used amino acids 25 to 50, for fungal sequences we used amino acids 35 to 80 and for oomycete sequences we used amino acids 15 to 40.
+Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs as shown in the logo plots in Fig \@ref(fig:activations) even within a taxon and data set.
+Enrichment analysis of different amino acid categories showed a statistically significant ($p < 0.05$) increase of hydrophobic amino acids in the effector sequences of bacteria and oomycete sequence regions relative to the combined effector and non effector set \@ref(tab:phyperaatype) and an enrichment of polar amino acids in oomycete sequences, though no enrichment of any type was found in the fungal effectors.
+These properties are consistent with the secreted role of known effector proteins.
 
@@
 The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
+
+<!--
+
+#########################################################
+## Start code fragments
+##
+##
+#########################################################
+
+-->
 
@@
 
+```{r, phyperaatype, echo=FALSE, message=FALSE}
+source(here::here("functions","check_proportions_aas.R"))
+bacteria_result <- join_files("data/raw/bacteria_testing.csv", "data/raw/bacteria_training.csv", "data/raw/bacteria_validation.csv") %>% 
+  truncate_seqs(start = 25, stop = 50) %>% 
+  do_hyper()
+fungi_result <- join_files("data/raw/fungi_testing.csv", "data/raw/fungi_training.csv", "data/raw/fungi_validation.csv") %>% 
+  truncate_seqs(start = 35, stop = 80) %>% 
+  do_hyper()
+oomycete_result <- join_files("data/raw/oomycete_testing.csv", "data/raw/oomycete_training.csv", "data/raw/oomycete_validation.csv") %>% 
+  truncate_seqs(start = 15, stop = 40) %>% 
+  do_hyper()
+
+r <- bind_rows(list(bacteria_result, fungi_result, oomycete_result))
+r$Taxon <- c("Bacteria", "Fungi", "Oomycete") 
+r <- r %>% select(Taxon, Hydrophobic, Polar, Neutral, Basic, Acidic)
+apa_table(
+  r,
+  digits = 3,
+  align = 'c',
+  caption = 'Enrichment analysis of amino acid types.
+Probability by hypergeometric test of the observed number of amino acid type in the region of activation in effector sequences in our compiled sequence set relative to a background of the same region in all sequences in that taxon.'
+)
+```
+
 # Discussion
@@
 # Supplementary Tables
+
+<!--
+
+#########################################################
+## Start supplementary information
+##
+##
+#########################################################
+
+-->
 

### 2020-06-08 functions/check_proportions_aas.R

@@ -1,0 +1,83 @@
+library(dplyr)
+library(tidyr)
+
+count_types <- function(seq) {
+  s <- strsplit(seq, split = "")[[1]]
+  tibble::tibble(
+  "hydrophobic" = sum(s %in% c("P", "A","W","F","L","I","M","V")) ,
+  "polar" = sum(s %in% c("G", "S", "T", "Y", "C" )),
+  "neutral" = sum(s %in% c("N", "Q")),
+  "basic" = sum(s %in% c("K","R","H")),
+  "acidic" = sum(s %in% c("D", "E")),
+  "length" = length(s)
+  )
+}
+
+
+get_counts <- function(df){
+  
+  df %>% 
+    mutate(res = purrr::map(Sequence,count_types)) %>% 
+    unnest(res)
+  
+}
+
+get_global_counts <- function(df){
+  
+  
+  counts <- get_counts(df)
+  
+  list(
+    total_length = sum(counts$length),
+    total_hydrophobic = sum(counts$hydrophobic),
+    total_polar = sum(counts$polar),
+    total_neutral = sum(counts$neutral),
+    total_basic = sum(counts$basic),
+    total_acidic = sum(counts$acidic)
+  )
+  
+}
+
+get_effector_counts <- function(df, keep_label = 1){
+  
+  counts <- df %>% filter(label == keep_label) %>% 
+    get_counts()
+  list(
+    total_length = sum(counts$length),
+    total_hydrophobic = sum(counts$hydrophobic),
+    total_polar = sum(counts$polar),
+    total_neutral = sum(counts$neutral),
+    total_basic = sum(counts$basic),
+    total_acidic = sum(counts$acidic)
+  )
+}
+
+
+join_files <- function(...){
+  files <- c(...)
+  seqs <- lapply(files, readr::read_csv)
+  names(seqs) <- files
+  bind_rows(seqs)
+}
+
+do_hyper <- function(df) {
+
+  all_seqs <- get_global_counts(df)
+  eff_seqs <- get_effector_counts(df)
+  tibble::tibble(
+    Hydrophobic = phyper(eff_seqs$total_hydrophobic, all_seqs$total_hydrophobic, (all_seqs$total_length - all_seqs$total_hydrophobic), eff_seqs$total_length ),
+    Polar = phyper(eff_seqs$total_polar, all_seqs$total_polar, (all_seqs$total_length - all_seqs$total_polar), eff_seqs$total_length ),
[…]

### 2020-06-08 main.Rmd

--- 2020-06-07 main.Rmd
@@
     latex_engine: pdflatex
-    # template: svm_latex.tex
+    template: svm_latex.tex
     toc: no
@@
 biblio-style: apsr
-bibliography: ["references.bib", "r-packages.bib"]
+bibliography: ["r-references.bib", "r-packages.bib"]
 graphics: yes
@@
 Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
-We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range $\sim 93.4 \%$  to  $\sim 97.4 \%$, for the four fungal models we observed accuracy in the range $\sim 60.5 \%$  to  $\sim 84.2 \%$  and for the four oomycete models we saw accuracy from $\sim 64.7 \%$  to  $\sim 82.3 \%$, reported in Fig \@ref(fig:overallmodel).
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel).
 All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
@@
 
-Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range \numrange[range-phrase = --]{0.80}{0.88}, see Figure \@ref(fig:confusionmodelcorrs).
+Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs).
 CNN-GRU and LSTM-Embedding showed identical prediction sets.
-We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range \numrange[range-phrase = --]{0.79}{0.90}; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range \numrange[range-phrase = --]{0.36}{0.51}.
-The models trained on oomycete sequences all showed this lower range of correlations, in the range \numrange[range-phrase = --]{0.30}{0.65}.
+We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51.
+The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65.
 The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves.
@@
 This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
-Absoloute improvements were up to $\sim 15\%$  in bacterial sequences, $\sim 15\%$ in fungal sequences and $\sim 20\%$  in the oomycete sequences.
+Absoloute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences.
 The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
@@
 A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
-To examine the responses of the model, we ran the effector sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the Grad-CAM method.
+To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
 The profiles were smoothed using FFT and examined \@ref(fig:activations).
@@
 To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
-The package ans installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff).
+The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
+
+<!--
+
+#########################################################
+## Start code fragments
+##
+##
+#########################################################
+
+-->
 
@@
 ```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
+
 # Load data
@@
 
-```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
+```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", echo=FALSE, message=FALSE}
 data.table::fread("data/processed/acc_f1score_vs_othertools.csv") %>%
@@
 ```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
-A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
-# knitr::include_graphics("figures/cf_other_tools.png")
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
 
@@
 
-```{r, bacefftable}
+<!--
+
+#########################################################
+## Start supplementary information
+##
+##
+#########################################################
+
+-->
+
+```{r, bacefftable, echo=FALSE, message = FALSE}
+
 apa_table(

### 2020-06-08 main.Rmd
Fixed #7 and #8

@@
 biblio-style: apsr
-bibliography: ["r-references.bib", "r-packages.bib"]
+bibliography: ["r-references.bib"]
 graphics: yes
@@
 # Build R pacakges Bib build
-r_refs(file = "r-packages.bib")
+papaja::r_refs(file = "r-references.bib")
 
@@
   labs(x = 'Position', y = "Value") +
-  ggplot2::labs(tag = "B")
+  ggplot2::labs(tag = "C")
 
@@
   labs(x = 'Position', y = "Value") +
-  ggplot2::labs(tag = "C")
+  ggplot2::labs(tag = "B")
 

### 2020-06-08 main.Rmd
allow tex files in commits

--- 2020-06-08 main.Rmd
@@
     toc: no
+    includes:
+      in_header: preamble.tex
 fontsize: 12pt
@@
 
-Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
+Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
 The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used.
@@
 We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).
-
-```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
-knitr::include_graphics(
-  "figures/flowchart_effectorsequence_collection.pdf"
-)
-```
-
-```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
-knitr::include_graphics(
-  "figures/flowchart_secreted_noneffectorsequence_collection.pdf"
-)
-```
 
@@
 -->
+```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+knitr::include_graphics(
+  "figures/flowchart_effectorsequence_collection.pdf"
+)
+```
+
+```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+knitr::include_graphics(
+  "figures/flowchart_secreted_noneffectorsequence_collection.pdf"
+)
+```
 
@@
 
+\beginsupplement
 ```{r, bacefftable, echo=FALSE, message = FALSE}
@@
 
-```{r hyperparam}
+```{r, hyperparam}
  readr::read_tsv("tables/hyperparamscan_info.tab") %>%

### 2020-06-08 main.Rmd
Move figures inline. Add long tables to supplement

@@
 
-## Model Selection and Training
-
-We selected four base architectures on which to build models for learning.
-Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU.
-All models had fully-connected dense layers after this.
-See Figure \@ref(fig:modelarch).
-Hyperparameter values for the models were identified using a random search as described in methods.
-Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
-For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.
-Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), a total of twelve models were produced, one for each taxon and model combination.
-The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).
-
-We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality.
-Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size.
-We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.
-Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
-We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel).
-All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
-
-## Model Characteristics
-
-We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.
-The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors.
-The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors.
-The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct.
-Overall the models are accurate and show little to no bias toward false positive or false negatives, with the exception of the GRU-Embedding type.
-In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.
-
-Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs).
-CNN-GRU and LSTM-Embedding showed identical prediction sets.
-We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51.
-The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65.
-The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves.
-For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others.
-The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set.
-It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.
-
-## Ensemble Models
-
-We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input.
-We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults).
-With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
-However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures.
-Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models.
-This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.
-
-## Comparison With Other Classification Software
-
-Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software.
-We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR].
-Each comparison was carried out using the respective hold-out test sequence set for each taxon.
-For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics).
-This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
-Absoloute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences.
-The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
-All the established software classifiers we tested show lower sensitivity than the models we have developed here.
-
-## Convolution Heatmaps
-
-An advantage of CNNs relative to other deep neural networks is their relative interpretability.
-A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
-To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
-The profiles were smoothed using FFT and examined \@ref(fig:activations).
-All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals.
-The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids.
-We examined further the sequences under the largest peaks, specifically for bacterial sequences we used amino acids 25 to 50, for fungal sequences we used amino acids 35 to 80 and for oomycete sequences we used amino acids 15 to 40.
-Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs as shown in the logo plots in Fig \@ref(fig:activations) even within a taxon and data set.
-Enrichment analysis of different amino acid categories showed a statistically significant ($p < 0.05$) increase of hydrophobic amino acids in the effector sequences of bacteria and oomycete sequence regions relative to the combined effector and non effector set \@ref(tab:phyperaatype) and an enrichment of polar amino acids in oomycete sequences, though no enrichment of any type was found in the fungal effectors.
-These properties are consistent with the secreted role of known effector proteins.
-
-## Software Implementation
-
-To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
-The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
-
-<!--
-
-#########################################################
-## Start code fragments
-##
-##
-#########################################################
-
--->
-```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+```{r effectorcollection, fig.width=10, fig.height=10, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
 knitr::include_graphics(
@@
 
-```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+```{r noneffectorcollection, fig.width=10, fig.height=10, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
 knitr::include_graphics(
@@
 
-```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model"}
+## Model Selection and Training
+
+We selected four base architectures on which to build models for learning.
+Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU.
+All models had fully-connected dense layers after this.
+See Figure \@ref(fig:modelarch).
+Hyperparameter values for the models were identified using a random search as described in methods.
+Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
+For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.
+Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), a total of twelve models were produced, one for each taxon and model combination.
+The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).
+
+```{r modelarch, fig.width=10, fig.height=10, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model"}
 knitr::include_graphics("figures/model_arch.pdf")
+
 ```
+
+```{r acclossfinal}
+ readr::read_tsv("tables/acc_loss_final.tab") %>%
+  knitr::kable( "latex", longtable = T, booktabs = T,
+           digits = 4,
+  align = 'c',
+  caption = 'Accuracy and loss values from best performing parameters values for each model'
+         ) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
+```
+
+We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality.
+Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size.
+We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.
+Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel).
+All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
 
@@
 ```
+
+## Model Characteristics
+
+We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.
+The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors.
+The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors.
+The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct.
+Overall the models are accurate and show little to no bias toward false positive or false negatives, with the exception of the GRU-Embedding type.
+In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.
+
+Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs).
+CNN-GRU and LSTM-Embedding showed identical prediction sets.
+We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51.
+The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65.
+The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves.
+For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others.
+The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set.
+It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.
 
@@
 
-```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+## Ensemble Models
+
+We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input.
+We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults).
+With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
+However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures.
+Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models.
+This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.
+
+```{r ensembleresults, fig.width=8, fig.height=8, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
 # apa_table(
@@
 
-```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", echo=FALSE, message=FALSE}
+## Comparison With Other Classification Software
+
+Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software.
+We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR].
+Each comparison was carried out using the respective hold-out test sequence set for each taxon.
+For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics).
+This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
+Absoloute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences.
+The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
+All the established software classifiers we tested show lower sensitivity than the models we have developed here.
+
+```{r vsothersmetrics, fig.width=8, fig.height=8,fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", echo=FALSE, message=FALSE}
 data.table::fread("data/processed/acc_f1score_vs_othertools.csv") %>%
@@
 
-```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
+```{r vsotherscfmat, fig.width=8, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
 A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
@@
 
-```{r activations, fig.height=6, fig.width=10, fig.cap="Grad-CAM activation profiles of CNN-LSTM models trained on when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data and non-effector data.
+## Convolution Heatmaps
+
+An advantage of CNNs relative to other deep neural networks is their relative interpretability.
+A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
+To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
+The profiles were smoothed using FFT and examined \@ref(fig:activations).
+All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals.
+The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids.
+We examined further the sequences under the largest peaks, specifically for bacterial sequences we used amino acids 25 to 50, for fungal sequences we used amino acids 35 to 80 and for oomycete sequences we used amino acids 15 to 40.
+Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs as shown in the logo plots in Fig \@ref(fig:activations) even within a taxon and data set.
+Enrichment analysis of different amino acid categories showed a statistically significant ($p < 0.05$) increase of hydrophobic amino acids in the effector sequences of bacteria and oomycete sequence regions relative to the combined effector and non effector set \@ref(tab:phyperaatype) and an enrichment of polar amino acids in oomycete sequences, though no enrichment of any type was found in the fungal effectors.
+These properties are consistent with the secreted role of known effector proteins.
+
+```{r activations, fig.height=10, fig.width=10, fig.cap="Grad-CAM activation profiles of CNN-LSTM models trained on when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data and non-effector data.
 A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
@@
 
+## Software Implementation
+
+To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
+The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
+
+<!--
+
+#########################################################
+## Start code fragments
+##
+##
+#########################################################
+
+-->
+
 # Discussion
@@
 ```
-
-```{r acclossfinal}
- readr::read_tsv("tables/acc_loss_final.tab") %>%
-  knitr::kable( "latex", longtable = T, booktabs = T,
-           digits = 4,
-  align = 'c',
-  caption = 'Accuracy and loss values from best performing parameters values for each model'
-         ) %>%
-  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
-```

### 2020-06-09 data/processed/compare_result_effectorp_data_prediction.R
Added R script to get acc, f1 score, and conf mat data from the effectorP data prediction results

@@ -1,0 +1,115 @@
+library(dplyr)
+
+# Function to get true and pred value from effectorp output
+get_true_pred_value_from_effectorp_pred <- function(data_path, label_data){
+  df <- data.table::fread(data_path) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(Prediction == "Effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector")), 
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  return(df)
+}
+
+# Function to get true and pred from deepredeff output
+get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
+  df <- data.table::fread(data_path) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  return(df)
+}
+
+# Get confusion matrix from both
+get_confusion_mat <- function(data_frame, label_data, data_origin, tool_used){
+  
+  if (tool_used != "cnn_lstm"){
+    df <- get_true_pred_value_from_effectorp_pred(data_frame, label_data = label_data)
+  } else {
+    df <- get_true_pred_value_from_deepredeff(data_frame, label_data = label_data) 
+  }
+  
+  conf_mat_df <- df %>%
+    yardstick::conf_mat(truth = label, estimate = pred) %>% 
+    .$table %>% 
+    as.data.frame() %>% 
+    filter(Truth != "non-effector") %>% 
+    mutate(
+      data = data_origin, 
+      tool = tool_used
+    )
+  
+  return(conf_mat_df)
+}
+
+# Get all accuracy and f1_score
+metrics_list <- yardstick::metric_set(
+  yardstick::accuracy,
+  yardstick::f_meas
+)
+
+
+# Calculate accuracy and f1_score
+acc_f1_score_effectorp_data_prediction <- rbind(
+  rbind(
+    get_true_pred_value_from_deepredeff("data/raw/effectorP1_pos_training_evaluated_on_deepredeff.csv", label_data = "effector") %>%
+      mutate(data = "effectorP1_training", tool = "cnn_lstm"), 
[…]

### 2020-06-09 main.Rmd
Added classification results on EffectorP data

@@
 
+We are also interested in evaluating positive datasets from other tools on our model and compare the results when the datasets evaluated on their own tool.
+Table \@ref(tab:accf1scoreeffectorpdata) shows the of accuracy and F1 score values of classification results of three different positive datasets from EffectorP 1.0 and 2.0.
+
+```{r accf1scoreeffectorpdata}
+apa_table(
+  readr::read_csv("data/processed/acc_f1score_result_effectorp_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of EffectorP data evaluated on Deepredeff and both EffectorP 1.0 and EffectorP 2.0',
+  placement = "h"
+)
+```
+
 ```{r vsothersmetrics, fig.width=8, fig.height=8,fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", echo=FALSE, message=FALSE}
@@
   plot_annotation(tag_levels = "A")
+```
+
+```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", message = FALSE, echo=FALSE}
+readr::read_csv("data/processed/conf_mat_effectorp_data_prediction.csv") %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(x = Truth, y = Prediction) +
+  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
+  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
+  ggplot2::facet_grid(data ~ tool) +
+  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
+  ggplot2::labs(x = "True value", y = "Prediction") +
+  ggplot2::coord_fixed()
 ```
@@
 \beginsupplement
+
 ```{r, bacefftable, echo=FALSE, message = FALSE}
-
 apa_table(
@@
   align = 'c',
-  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
+  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base',
+  placement = "h"
 )
@@
 readr::read_tsv("tables/hyperparamscan_info.tab") %>%
-  knitr::kable( "latex", longtable = T, booktabs = T,
+  knitr::kable(
+    "latex", 
+    longtable = T, 
+    booktabs = T,
     digits = 0,
@@
   kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
-
 ```

### 2020-06-09 compare_result_effectivet3_data_prediction.R
Added R script to process the result to process the classification result

@@ -1,0 +1,97 @@
+library(dplyr)
+
+# Function to get true and pred value from effectorp output
+get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
+  df <- data.table::fread(data_path) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  return(df)
+}
+
+
+get_true_pred_value_from_effectivet3 <- function(data_path, label_data){
+  
+  df <- data.table::fread(data_path, skip = 1) %>% 
+    dplyr::select(-c("Description", "Score")) %>% 
+    `colnames<-`(c("id", "is_effective")) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(is_effective == TRUE, 1, 0), levels = c(1, 0), labels = c("effector", "non-effector")), 
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  
+  return(df)
+}
+
+
+# Get confusion matrix from both
+get_confusion_mat <- function(data_path, label_data, data_origin, tool_used){
+  
+  if (tool_used != "ensemble_weighted"){
+    df <- get_true_pred_value_from_effectivet3(data_path, label_data = label_data)
+  } else {
+    df <- get_true_pred_value_from_deepredeff(data_path, label_data = label_data) 
+  }
+  
+  conf_mat_df <- df %>%
+    yardstick::conf_mat(truth = label, estimate = pred) %>% 
+    .$table %>% 
+    as.data.frame() %>% 
+    filter(Truth != "non-effector") %>% 
+    mutate(
+      data = data_origin, 
+      tool = tool_used
+    )
+  
+  return(conf_mat_df)
+}
+
+
+
+# Get all accuracy and f1_score
+metrics_list <- yardstick::metric_set(
+  yardstick::accuracy,
+  yardstick::f_meas
+)
+
+
+acc_f1_score_effectivet3_data_prediction <- rbind(
+rbind(
[…]

### 2020-06-09 main.Rmd
Added table of acc and f1score, plot of conf_mat in main.Rmd

@@
   - \usepackage{siunitx}
+  - \usepackage{float}
 ---
@@
 
-```{r ensembleresults, fig.width=8, fig.height=8, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r ensembleresults, fig.width=8, fig.height=4, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
 # apa_table(
@@
   align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of EffectorP data evaluated on Deepredeff and both EffectorP 1.0 and EffectorP 2.0',
-  placement = "h"
+  caption = 'Accuracy and F1-Score from prediction results of EffectorP data evaluated on our model and both EffectorP 1.0 and EffectorP 2.0',
+  placement = "H"
 )
@@
 
-```{r vsothersmetrics, fig.width=8, fig.height=8,fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", echo=FALSE, message=FALSE}
+```{r accf1scoreeffectivet3data}
+apa_table(
+  readr::read_csv("data/processed/acc_f1score_result_effectivet3_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of EffeciveT3 positive data evaluated on our model and EffectiveT3 trained on both plant associated protein and all protein.',
+  placement = "H"
+)
+```
+
+```{r vsothersmetrics, fig.width=8, fig.height=4, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", fig.pos="H"}
 data.table::fread("data/processed/acc_f1score_vs_othertools.csv") %>%
@@
 ```{r vsotherscfmat, fig.width=8, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
-A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", fig.pos="H"}
 
@@
 
-```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", message = FALSE, echo=FALSE}
+```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", fig.pos="H"}
+
 readr::read_csv("data/processed/conf_mat_effectorp_data_prediction.csv") %>% 
@@
   ggplot2::facet_grid(data ~ tool) +
+  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
+  ggplot2::labs(x = "True value", y = "Prediction") +
+  ggplot2::coord_fixed()
+```
+
+```{r effectivet3datacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H"}
+
+readr::read_csv("data/processed/conf_mat_effectivet3_data_prediction.csv") %>% 
+  dplyr::mutate(
+  tool = factor(tool, levels = c("ensemble_weighted", "effectivet3_all", "effectivet3_plant"))
+  ) %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(x = Truth, y = Prediction) +
+  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
+  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
+  ggplot2::facet_grid(~tool) +
   ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +

### 2020-06-09 data/processed/compare_result_deept3_data_prediction.R
Added R script to process the prediction result

@@ -1,0 +1,102 @@
+library(dplyr)
+
+# Function to get true and pred from deepredeff output
+get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
+  df <- data.table::fread(data_path) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  return(df)
+}
+
+# Function to get true and pred value from effectorp output
+get_true_pred_value_from_deept3_pred <- function(data_path, label_data){
+  df <- data.table::fread(data_path, header = FALSE) %>% 
+    `colnames<-`(c("Prediction")) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(Prediction == "T3SE", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector")), 
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  return(df)
+}
+
+# Get confusion matrix from both
+get_confusion_mat <- function(data_path, label_data, data_origin, tool_used){
+  
+  if (tool_used != "ensemble_weighted"){
+    df <- get_true_pred_value_from_deept3_pred(data_path, label_data = label_data)
+  } else {
+    df <- get_true_pred_value_from_deepredeff(data_path, label_data = label_data) 
+  }
+  
+  conf_mat_df <- df %>%
+    yardstick::conf_mat(truth = label, estimate = pred) %>% 
+    .$table %>% 
+    as.data.frame() %>% 
+    filter(Truth != "non-effector") %>% 
+    mutate(
+      data = data_origin, 
+      tool = tool_used
+    )
+  
+  return(conf_mat_df)
+}
+
+# Get all accuracy and f1_score
+metrics_list <- yardstick::metric_set(
+  yardstick::accuracy,
+  yardstick::f_meas
+)
+
+# Calculate accuracy and f1_score
+acc_f1_score_deept3_data_prediction <- rbind(
+rbind(
+  get_true_pred_value_from_deepredeff("data/raw/deept3_P.syringae_nr_effector_data_evaluated_on_deepredeff.csv", label_data = "effector") %>% 
[…]

### 2020-06-09 main.Rmd
Added acc and f1_score table, conf_mat plot of DeepT3 data classification result to main.Rmd

@@
 
+```{r accf1scoredeept3data}
+apa_table(
+  readr::read_csv("data/processed/acc_f1score_result_deept3_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of three different DeepT3 positive datasets evaluated on our model and DeepT3.',
+  placement = "H"
+)
+```
+
 ```{r vsothersmetrics, fig.width=8, fig.height=4, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", fig.pos="H"}
@@
   ggplot2::facet_grid(~tool) +
+  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
+  ggplot2::labs(x = "True value", y = "Prediction") +
+  ggplot2::coord_fixed()
+```
+
+```{r deept3datacfmat, fig.height=8, fig.cap="Confusion matrices of classifications made by our bacteria model compared with DeepT3 evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
+
+readr::read_csv("data/processed/conf_mat_deept3_data_prediction.csv") %>% 
+  dplyr::mutate(
+  tool = factor(tool, levels = c("ensemble_weighted", "deept3"))
+  ) %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(x = Truth, y = Prediction) +
+  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
+  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
+  ggplot2::facet_grid(data ~ tool) +
   ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +

### 2020-06-10 data/processed/compare_result_effectR_data_prediction.R
Added r-script to process the raw result

@@ -1,0 +1,97 @@
+library(dplyr)
+
+# Function to get true and pred from deepredeff output
+get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
+  df <- data.table::fread(data_path) %>% 
+    dplyr::mutate(
+      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
+      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
+    )
+  return(df)
+}
+
+get_confusion_mat <- function(data_frame, filter){
+  
+  
+  conf_mat_df <- data_frame %>%
+    dplyr::filter(tool == filter) %>%
+    yardstick::conf_mat(truth = label, estimate = value) %>% 
+    .$table %>% 
+    as.data.frame() %>% 
+    filter(Truth != "non-effector") %>% 
+    mutate(
+      tool = filter
+    )
+  
+  return(conf_mat_df)
+}
+
+
+# Get all accuracy and f1_score
+metrics_list <- yardstick::metric_set(
+  yardstick::accuracy,
+  yardstick::f_meas
+)
+
+
+
+effectR_data_evaluated_on_deepredeff_and_effectR <- get_true_pred_value_from_deepredeff(
+  "data/raw/effectR_reference_rxlr_effectors_evaluated_on_deepredeff.csv", 
+  label_data = "effector"
+) %>% 
+  mutate(
+    name = sub(" .*", "", name)
+  ) %>% 
+  select(name, label, pred_cnn_lstm = pred) %>% 
+  left_join(
+    data.table::fread("data/raw/effectR_reference_rxlr_effectors_evaluated_on_effectR.csv") %>% 
+      select(name, pred_effectR = candidate), 
+    by = "name"
+  ) %>% 
+  dplyr::mutate(
+    pred_effectR = ifelse(is.na(pred_effectR), "non-effector", "effector")
+  ) %>% 
+  tidyr::pivot_longer(
+    cols = -c(name, label),
+    names_to = "tool"
+  ) %>% 
+  dplyr::mutate(
+    value = factor(value, levels = c("effector", "non-effector")),
+    label = factor(label, levels = c("effector", "non-effector"))
+  ) 
+
[…]

### 2020-06-10 main.Rmd
Added the acc, f1-score table and plot of conf_mat from classification result using effectR reference sequences in main.Rmd

@@
 
+```{r accf1scoreeffectRdata}
+apa_table(
+  readr::read_csv("data/processed/acc_f1_score_effectR_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of reference sequences of EffectR evaluated on our model and EffectR.',
+  placement = "H"
+)
+```
+
 ```{r vsothersmetrics, fig.width=8, fig.height=4, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", fig.pos="H"}
@@
 
-```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", fig.pos="H"}
+```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", fig.pos="H", message=FALSE}
 
@@
 
-```{r effectivet3datacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H"}
+```{r effectivet3datacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
 
@@
   ggplot2::facet_grid(data ~ tool) +
+  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
+  ggplot2::labs(x = "True value", y = "Prediction") +
+  ggplot2::coord_fixed()
+```
+
+```{r effectRdatacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
+
+readr::read_csv("data/processed/conf_mat_effectR_data_prediction.csv") %>% 
+  dplyr::mutate(
+  tool = factor(tool, levels = c("cnn_lstm", "effectR"))
+  ) %>% 
+  ggplot2::ggplot() +
+  ggplot2::aes(x = Truth, y = Prediction) +
+  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
+  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
+  ggplot2::facet_grid(~tool) +
   ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +

### 2020-06-10 main.Rmd
added some text on comparison woth other models

@@
 Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software.
-We compared against predictions from the bacterial effector prediction programs deepT3 [@deept3] and effectiveT3 [@effectivet3], the fungal effector prediction programs EffectorP 1 and 2 [@effectorp, @effectorp2] and the oomycete effector predictor effectR [@effectR].
+We compared against predictions from the bacterial effector prediction programs DeepT3 [@deept3] and EffectiveT3 classification module for plant-associated bacteria 1.0.1 [@effectivet3], the fungal effector prediction programs EffectorP 1.0 and 2.0 [@effectorp, @effectorp2] and the oomycete effector predictor EffectR [@effectR].
 Each comparison was carried out using the respective hold-out test sequence set for each taxon.
@@
 
-We are also interested in evaluating positive datasets from other tools on our model and compare the results when the datasets evaluated on their own tool.
-Table \@ref(tab:accf1scoreeffectorpdata) shows the of accuracy and F1 score values of classification results of three different positive datasets from EffectorP 1.0 and 2.0.
+We are also interested in evaluating positive datasets from other tools on both our model and on theirs, to be able to have a more unbiased comparison of the models' performance.
+Table \@ref(tab:accf1scoreeffectorpdata) shows the of accuracy and F1 score values of classification results of three different positive datasets from EffectorP 1.0 and 2.0 which can be found in [http://effectorp.csiro.au./data.html](http://effectorp.csiro.au./data.html).
+As it can be seen, EffectorP tends perform better on their training data compared to the validation data, which indicates that EffectorP is likely overfitted to the traning data which makes it less generalised.
+
+The accuracy and F1 score values of evaluating positive dataset of EffectiveT3 obtained from [https://effectors.csb.univie.ac.at/sites/eff/files/others/TTSS_positive_training.faa](https://effectors.csb.univie.ac.at/sites/eff/files/others/TTSS_positive_training.faa) and positive datasets of DeepT3 obtained from [https://github.com/lje00006/DeepT3/tree/master/DeepT3/DeepT3-Keras/data](https://github.com/lje00006/DeepT3/tree/master/DeepT3/DeepT3-Keras/data) are shown in Table
+\@ref(tab:accf1scoreeffectivet3data) and  \@ref(tab:accf1scoredeept3data), respectively.
+Our ensemble weighted bacteria model performs better in classifing effector than using both EffectiveT3 for plant-associated bacteria and for all bacteria (see Table \@ref(tab:accf1scoreeffectivet3data)).
+Similar case with DeepT3 datasets, overall as shown in Table \@ref(tab:accf1scoredeept3data), our ensemble model for bacteria improved the performance of DeepT3.
+
+Table \@ref(tab:accf1scoreeffectRdata) shows the results of examining several reference RxLR effectors (PexRD36, PexRD1, ipi01/Avrblb1, Avr1, Avr4, and Avr3a) used by EffectR [@effectR] on both our CNN-LSTM model and EffectR.
+It is shown that our model is able to recognise around 64% of the oomycete effector sequence data without having any prior knowledge of effector motifs as it is required in EffectR.
 
@@
 
-```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", fig.pos="H", message=FALSE}
+```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", fig.pos="H", message=FALSE, include=FALSE}
 
@@
 
-```{r effectivet3datacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
+```{r effectivet3datacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE, include=FALSE}
 
@@
 
-```{r deept3datacfmat, fig.height=8, fig.cap="Confusion matrices of classifications made by our bacteria model compared with DeepT3 evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
+```{r deept3datacfmat, fig.height=8, fig.cap="Confusion matrices of classifications made by our bacteria model compared with DeepT3 evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE, include=FALSE}
 
@@
 
-```{r effectRdatacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
+```{r effectRdatacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE, include=FALSE}
 
@@
 A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
-To examine the responses of the model, we ran the effector sequence ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another] data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
+To examine the responses of the model, we ran all sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
 The profiles were smoothed using FFT and examined \@ref(fig:activations).

### 2020-06-10 main.Rmd
add figure and text for comparisons and method

--- 2020-06-09 main.Rmd
@@
 The code on computing and visualising these heatmaps can be found on [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .
+
+## Effector Prediction Software and Training data
+
+To test the performance of our models against commonly used tools we used DeepT3 version 1 [@deept3] and EffectiveT3 version 1.0.1 [@effectivet3] for bacterial sequences.
+Effector P Version 1.0 and 2.0 [@effectorp, @effectorp2] for fungal sequences and EffectR [@effectR] for oomycete sequences.
+All the models publish the datasets used to train the models or some examples.
+We used positive training examples (effector sequence examples) in the comparisons we performed.
+EffectorP provides three different positive datasets (training, test and hold-out validation) for EffectorP 1.0 and 2.0 at [http://effectorp.csiro.au./data.html](http://effectorp.csiro.au./data.html).
+EffectiveT3 provides a training set at [https://effectors.csb.univie.ac.at/sites/eff/files/others/TTSS_positive_training.faa](https://effectors.csb.univie.ac.at/sites/eff/files/others/TTSS_positive_training.faa).
+DeepT3 provides three sets, a non redundant Pseudomonas syringae effector dataset, a training dataset and a test data set at [https://github.com/lje00006/DeepT3/tree/master/DeepT3/DeepT3-Keras/data](https://github.com/lje00006/DeepT3/tree/master/DeepT3/DeepT3-Keras/data).
+EffectR provides 6 oomycete sequences, namely *PexRD36*, *PexRD1*, *ipi01/Avrblb1*, *Avr1*, *Avr4*, and *Avr3a*.
+All these sets were used in their respective tools with default settings.
 
@@
 This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
-Absoloute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences.
+Absolute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences.
 The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
@@
 
-We are also interested in evaluating positive datasets from other tools on our model and compare the results when the datasets evaluated on their own tool.
-Table \@ref(tab:accf1scoreeffectorpdata) shows the of accuracy and F1 score values of classification results of three different positive datasets from EffectorP 1.0 and 2.0.
+We also evaluated the deep learning models we have developed on the training and hold-out validation sequences used to train the previous methods.
+We calculated accuracy and F1 score of all the tools on their respective training and validation sets, according to availability.
+The bacterial tools EffectiveT3 and DeepT3 showed lower accuracies and F1 score on all sequence sets tested, than our Ensemble Weighted model, as in Figure \@ref(fig:owndata)A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated.
+Interestingly, both versions of EffectorP showed greater accuracy and F1 score on the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically in both metrics \@ref(fig:owndata)B.
+The EffectorP 1 and 2 scores on validation data are well below the scores for the training data, a result that is usually interpreted as being evidence of an overfitted and less generalisable model.
+Our CNN-LSTM model for fungi showed similar scores across the training and validation set indicating a greater generability.
+Only six effector sequences are provided as examples with the oomycete specific effector finder effectR.
+As this is not a trainable model in the same sense as the others, no large training set is needed.
+We attempted classification with these and effectR was able to classify X of the six ^[Ruth, can you add the actual numbers of classified for this one], whereas our CNN-LSTM model for oomycetes classified  Y of the six\ref(fig:owndata)C.
 
-```{r accf1scoreeffectorpdata}
-apa_table(
-  readr::read_csv("data/processed/acc_f1score_result_effectorp_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of EffectorP data evaluated on our model and both EffectorP 1.0 and EffectorP 2.0',
-  placement = "H"
-)
-```
+Overall our model performed more strongly than those previously available ones tested across the range of sequences examined.
 
-```{r accf1scoreeffectivet3data}
-apa_table(
-  readr::read_csv("data/processed/acc_f1score_result_effectivet3_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of EffeciveT3 positive data evaluated on our model and EffectiveT3 trained on both plant associated protein and all protein.',
-  placement = "H"
-)
-```
-
-```{r accf1scoredeept3data}
-apa_table(
-  readr::read_csv("data/processed/acc_f1score_result_deept3_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of three different DeepT3 positive datasets evaluated on our model and DeepT3.',
-  placement = "H"
-)
-```
-
-```{r vsothersmetrics, fig.width=8, fig.height=4, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", fig.pos="H"}
+```{r vsothersmetrics, fig.width=8, fig.height=6, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", fig.pos="H"}
 data.table::fread("data/processed/acc_f1score_vs_othertools.csv") %>%
@@
 
-```{r effectorpdatacfmat, fig.width=6, fig.height=6, fig.cap="Confusion matrices of classifications made by our fungi model compared with EffectorP 1.0 and EffectorP 2.0 evaluated on both training and validation EffectorP data.", fig.pos="H"}
+```{r, owndata, fig.height = 8, fig.cap="Performance of models on training and validation datasets used to train other effector prediction tools.
+A: Tools trained on bacterial sequences; B: Tools trained on fungal sequences; C: Tools trained on oomycete sequences."}
+d <- readr::read_csv(here::here("data","processed", "owndatasets.csv"))
 
-readr::read_csv("data/processed/conf_mat_effectorp_data_prediction.csv") %>% 
-  ggplot2::ggplot() +
-  ggplot2::aes(x = Truth, y = Prediction) +
-  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
-  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
-  ggplot2::facet_grid(data ~ tool) +
-  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
-  ggplot2::labs(x = "True value", y = "Prediction") +
-  ggplot2::coord_fixed()
-```
+  mkplot <- function(df, which_tax ) {
+    df$Tool <- factor(df$Tool, levels = c("cnn_lstm", "ens_weight", "Effective T3 All", "Effective T3 Plants", "Deep T3", "EffectorP", "EffectorP 2", "effectR"))
+    p <- df %>% filter(Taxon == which_tax) %>% 
+     ggplot() +
+    aes(Tool, Value) +
+    geom_point(aes(colour = Metric)) +
+    facet_grid(  ~ `Data Set`, scales = "free") + 
+      theme_bw() +
+      scale_color_viridis_d()
+  }
 
-```{r effectivet3datacfmat, fig.width=6, fig.height=3, fig.cap="Confusion matrices of classifications made by our bacteria model compared with both EffectiveT3 based on plant associated and all protein evaluated on positive EffectiveT3 data.", fig.pos="H"}
+bac <- mkplot(d, "Bacteria")  + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.title.x = element_blank()) 
+fungi <- mkplot(d, "Fungi") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.title.x = element_blank())
+#fungi <- mkplot(d, "Fungi")
+oom <- mkplot(d, "Oomycete")
 
-readr::read_csv("data/processed/conf_mat_effectivet3_data_prediction.csv") %>% 
-  dplyr::mutate(
-  tool = factor(tool, levels = c("ensemble_weighted", "effectivet3_all", "effectivet3_plant"))
-  ) %>% 
-  ggplot2::ggplot() +
-  ggplot2::aes(x = Truth, y = Prediction) +
-  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
-  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
-  ggplot2::facet_grid(~tool) +
-  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
-  ggplot2::labs(x = "True value", y = "Prediction") +
-  ggplot2::coord_fixed()
-```
-
-```{r deept3datacfmat, fig.height=8, fig.cap="Confusion matrices of classifications made by our bacteria model compared with DeepT3 evaluated on positive EffectiveT3 data.", fig.pos="H", message=FALSE}
-
-readr::read_csv("data/processed/conf_mat_deept3_data_prediction.csv") %>% 
-  dplyr::mutate(
-  tool = factor(tool, levels = c("ensemble_weighted", "deept3"))
-  ) %>% 
-  ggplot2::ggplot() +
-  ggplot2::aes(x = Truth, y = Prediction) +
-  ggplot2::geom_tile(ggplot2::aes(fill = Freq), colour = "white") +
-  ggplot2::geom_text(ggplot2::aes(label = Freq), vjust = 0.5) +
-  ggplot2::facet_grid(data ~ tool) +
-  ggplot2::scale_fill_viridis_c(begin = 0.2, end = 1) +
-  ggplot2::labs(x = "True value", y = "Prediction") +
-  ggplot2::coord_fixed()
+#bac_eft3 #| 
+bac / fungi / oom + plot_layout(guides = "collect") + plot_annotation(tag_levels = 'A')
 ```
@@
 ```
+
+```{r accf1scoreeffectorpdata}
+apa_table(
+  readr::read_csv("data/processed/acc_f1score_result_effectorp_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of EffectorP data evaluated on our model and both EffectorP 1.0 and EffectorP 2.0',
+  placement = "H"
+)
+```
+
+```{r accf1scoreeffectivet3data}
+apa_table(
+  readr::read_csv("data/processed/acc_f1score_result_effectivet3_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of EffeciveT3 positive data evaluated on our model and EffectiveT3 trained on both plant associated protein and all protein.',
+  placement = "H"
+)
+```
+
+```{r accf1scoredeept3data}
+apa_table(
+  readr::read_csv("data/processed/acc_f1score_result_deept3_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of three different DeepT3 positive datasets evaluated on our model and DeepT3.',
+  placement = "H"
+)
+```
+
+```{r accf1scoreeffectRdata}
+apa_table(
+  readr::read_csv("data/processed/acc_f1_score_effectR_data_prediction.csv"),
+  digits = 4,
+  align = 'c',
+  caption = 'Accuracy and F1-Score from prediction results of reference sequences of EffectR evaluated on our model and EffectR.',
+  placement = "H"
+)
+```

### 2020-06-11 main.Rmd
edited the number EffectR data predicted by cnn-lstm and EffectR.

@@
 As this is not a trainable model in the same sense as the others, no large training set is needed.
-We attempted classification with these and effectR was able to classify X of the six ^[Ruth, can you add the actual numbers of classified for this one], whereas our CNN-LSTM model for oomycetes classified  Y of the six\ref(fig:owndata)C.
+We attempted classification with these and effectR was able to classify 12 of the 14, whereas our CNN-LSTM model for oomycetes classified nine of the 14, see Figure \@ref(fig:owndata)C.
 

### 2020-06-11 main.Rmd
remove unused files. Redo hypergeom

@@
 To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we adapted a Grad-CAM (Gradient-weighted Class Activation Mapping) approach [@Selvaraju2019].
-Using this method, we could produce the visualization of one of the convolutional neural network output values when it does a prediction.
-This will allow us to see which position in peptide sequences is important in determining whether it is effector or non-effector.
+We extracted the convolutional layer `convd_1` and computed the gradient of the model output with respect to the output of `convd_1` The feature map was weighted by every channel and mean result taken.
+All activation maps were summed, normalised and smoothed using Fourier analysis ^[Ruth, there need to be references, mentions of software and versions used, it isn't easy to see what you actually did from this description].
+The code we used for computing and visualising these heatmaps can be found at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .
 
-To find the importance of sequence position detected by our model, we took the output feature map of a convolution layer, given a sequence input.
-The convolutional layer we chose is the `convd_1` layer.
-Then, we computed the gradient of model output with the respect of the output of convolutional layer we chose.
-After that, we weighed every channel of that feature map with the gradient we calculated earlier, and took an average of the result.
-
-Following the process mentioned above, we then summed all of the activation map results of all sequences in our datasets for each taxon and normalised them.
-Next, the final summed activation map for each taxon was smoothed using fourier analysis and plotted the smoothed version for each of them.
-The code on computing and visualising these heatmaps can be found on [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .
-
-## Effector Prediction Software and Training data
+## Effector Prediction Software and Training Data
 
@@
 DeepT3 provides three sets, a non redundant Pseudomonas syringae effector dataset, a training dataset and a test data set at [https://github.com/lje00006/DeepT3/tree/master/DeepT3/DeepT3-Keras/data](https://github.com/lje00006/DeepT3/tree/master/DeepT3/DeepT3-Keras/data).
-EffectR provides 6 oomycete sequences, namely *PexRD36*, *PexRD1*, *ipi01/Avrblb1*, *Avr1*, *Avr4*, and *Avr3a*.
+EffectR uses 6 RXLR oomycete sequences described in [@haas2009genome] as examples rather than as a training set, namely *PexRD36*, *PexRD1*, *ipi01/Avrblb1*, *Avr1*, *Avr4*, and *Avr3a* .
 All these sets were used in their respective tools with default settings.
@@
 
-```{r acclossfinal}
+```{r, acclossfinal}
  readr::read_tsv("tables/acc_loss_final.tab") %>%
-  knitr::kable( "latex", longtable = T, booktabs = T,
-           digits = 4,
+  apa_table(            digits = 3,
   align = 'c',
   caption = 'Accuracy and loss values from best performing parameters values for each model'
-         ) %>%
-  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
+         )
 ```
@@
 We also evaluated the deep learning models we have developed on the training and hold-out validation sequences used to train the previous methods.
-We calculated accuracy and F1 score of all the tools on their respective training and validation sets, according to availability.
-The bacterial tools EffectiveT3 and DeepT3 showed lower accuracies and F1 score on all sequence sets tested, than our Ensemble Weighted model, as in Figure \@ref(fig:owndata)A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated.
-Interestingly, both versions of EffectorP showed greater accuracy and F1 score on the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically in both metrics \@ref(fig:owndata)B.
+We calculated the proportion of the effectors in the training set that the tools could find on their respective training and validation sets, according to availability.
+The bacterial tools EffectiveT3 and DeepT3 showed lower proportion found than our Ensemble Weighted model, as in Figure \@ref(fig:owndata)A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated.
+Interestingly, both versions of EffectorP found a greater proportion of the effectors in the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically (Figure \@ref(fig:owndata)B ).
 The EffectorP 1 and 2 scores on validation data are well below the scores for the training data, a result that is usually interpreted as being evidence of an overfitted and less generalisable model.
-Our CNN-LSTM model for fungi showed similar scores across the training and validation set indicating a greater generability.
+Our CNN-LSTM model for fungi showed similar scores across the training and validation set indicating a greater generability and equivalent power.
 Only six effector sequences are provided as examples with the oomycete specific effector finder effectR.
 As this is not a trainable model in the same sense as the others, no large training set is needed.
-We attempted classification with these and effectR was able to classify 12 of the 14, whereas our CNN-LSTM model for oomycetes classified nine of the 14, see Figure \@ref(fig:owndata)C.
+We attempted classification with these and effectR was able to classify 5 of the 6, whereas our CNN-LSTM model for oomycetes classified 3 of the 6 ( Figure \@ref(fig:owndata)C ).
 
-Overall our model performed more strongly than those previously available ones tested across the range of sequences examined.
+Overall, our models performed more strongly than the previously available ones tested across the range of sequences examined.
 
@@
      ggplot() +
-    aes(Tool, Value) +
-    geom_point(aes(colour = Metric)) +
+    aes(Tool, `Proportion Found`) +
+    geom_bar(stat= 'identity', fill = rgb(52/255, 106/255, 141/255), colour = rgb(52/255, 106/255, 141/255), alpha=0.5) +
     facet_grid(  ~ `Data Set`, scales = "free") + 
@@
 An advantage of CNNs relative to other deep neural networks is their relative interpretability.
-A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify.
-To examine the responses of the model, we ran all sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
-The profiles were smoothed using FFT and examined \@ref(fig:activations).
+A CNN can be analysed and activation maps extracted which highlight the regions of the input to which the CNN is most strongly relying on to classify.
+To examine the responses of the models, we ran all sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
+The profiles were smoothed using FFT and examined, (see Figure \@ref(fig:activations) ).
 All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals.
@@
 Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs as shown in the logo plots in Fig \@ref(fig:activations) even within a taxon and data set.
-Enrichment analysis of different amino acid categories showed a statistically significant ($p < 0.05$) increase of hydrophobic amino acids in the effector sequences of bacteria and oomycete sequence regions relative to the combined effector and non effector set \@ref(tab:phyperaatype) and an enrichment of polar amino acids in oomycete sequences, though no enrichment of any type was found in the fungal effectors.
-These properties are consistent with the secreted role of known effector proteins.
+Enrichment analysis of different amino acid categories showed some statistically significant ($p < 0.05$) changes in proportions of amino acid types relative to the whole set of sequences in the activated regions (Table \@ref(tab:phyperaatype) ).
+Bacterial effectors are depleted of hydrophobic amino acids and enriched in polar and basic amino acids.
+Fungal effectors are enriched in basic amino acids, with no other differences.
+The oomycete effectors are most interesting in that their activating regions are depleted in hydrophobic and polar amino acids while enriched in neutral and acidic amino acids.
 
-```{r activations, fig.height=10, fig.width=10, fig.cap="Grad-CAM activation profiles of CNN-LSTM models trained on when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data and non-effector data.
+```{r activations, fig.height=10, fig.width=10, fig.cap="Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data (left panel) and non-effector data (right panel).
 A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
@@
 r$Taxon <- c("Bacteria", "Fungi", "Oomycete") 
-r <- r %>% select(Taxon, Hydrophobic, Polar, Neutral, Basic, Acidic)
+#r <- r %>% select(Taxon, Hydrophobic, Polar, Neutral, Basic, Acidic)
+r <- r %>% 
+  mutate(Hydrophobic = paste0(sprintf("%.2f", prop_hydro), " : ", sprintf("%.2f",eff_prop_hydro), " (", sprintf("%.2f", p_hydro), ")"),
+         Polar = paste0(sprintf("%.2f", prop_polar), " : ", sprintf("%.2f ",eff_prop_polar), " (", sprintf("%.2f", p_polar), ")"),
+         Neutral = paste0(sprintf("%.2f",prop_neutral), " : ", sprintf("%.2f",eff_prop_neutral), " (", sprintf("%.2f",p_neutral), ")"),
+         Basic = paste0(sprintf("%.2f",prop_basic), " : ", sprintf("%.2f",eff_prop_basic), " (", sprintf("%.2f",p_basic), ")"),
+         Acidic = paste0(sprintf("%.2f",prop_acidic), " : ", sprintf("%.2f",eff_prop_acidic), " (", sprintf("%.2f",p_acidic), ")")
+         ) %>% select(Taxon, Hydrophobic, Polar, Neutral, Basic, Acidic)
 apa_table(
@@
   align = 'c',
+  font_size = 'footnotesize',
   caption = 'Enrichment analysis of amino acid types.
+Proportions of each amino acid type in the activation region of All Sequences : Effector Sequences, with (Probability).
 Probability by hypergeometric test of the observed number of amino acid type in the region of activation in effector sequences in our compiled sequence set relative to a background of the same region in all sequences in that taxon.'
 )
+
 ```
@@
 # Discussion
+
+We have compiled a set of known experimentally validated effectors from phytopathogenic bacteria, fungi and oomycetes and used them as positive training examples with which to tune a range of CNN classifiers.
 
@@
   align = 'c',
-  caption = 'Hyperparameters tuned in automatic scanning of hyperparameter space for each model'
+  caption = 'Hyperparameters tuned in manual scanning of hyperparameter space for each model'
 )
@@
 
-```{r accf1scoreeffectorpdata}
-apa_table(
-  readr::read_csv("data/processed/acc_f1score_result_effectorp_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of EffectorP data evaluated on our model and both EffectorP 1.0 and EffectorP 2.0',
-  placement = "H"
-)
 ```
-
-```{r accf1scoreeffectivet3data}
-apa_table(
-  readr::read_csv("data/processed/acc_f1score_result_effectivet3_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of EffeciveT3 positive data evaluated on our model and EffectiveT3 trained on both plant associated protein and all protein.',
-  placement = "H"
-)
-```
-
-```{r accf1scoredeept3data}
-apa_table(
-  readr::read_csv("data/processed/acc_f1score_result_deept3_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of three different DeepT3 positive datasets evaluated on our model and DeepT3.',
-  placement = "H"
-)
-```
-
-```{r accf1scoreeffectRdata}
-apa_table(
-  readr::read_csv("data/processed/acc_f1_score_effectR_data_prediction.csv"),
-  digits = 4,
-  align = 'c',
-  caption = 'Accuracy and F1-Score from prediction results of reference sequences of EffectR evaluated on our model and EffectR.',
-  placement = "H"
-)
-```

### 2020-06-11 data/processed/compare_result_deept3_data_prediction.R

@@ -1,102 +1,0 @@
-library(dplyr)
-
-# Function to get true and pred from deepredeff output
-get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
-  df <- data.table::fread(data_path) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  return(df)
-}
-
-# Function to get true and pred value from effectorp output
-get_true_pred_value_from_deept3_pred <- function(data_path, label_data){
-  df <- data.table::fread(data_path, header = FALSE) %>% 
-    `colnames<-`(c("Prediction")) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(Prediction == "T3SE", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector")), 
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  return(df)
-}
-
-# Get confusion matrix from both
-get_confusion_mat <- function(data_path, label_data, data_origin, tool_used){
-  
-  if (tool_used != "ensemble_weighted"){
-    df <- get_true_pred_value_from_deept3_pred(data_path, label_data = label_data)
-  } else {
-    df <- get_true_pred_value_from_deepredeff(data_path, label_data = label_data) 
-  }
-  
-  conf_mat_df <- df %>%
-    yardstick::conf_mat(truth = label, estimate = pred) %>% 
-    .$table %>% 
-    as.data.frame() %>% 
-    filter(Truth != "non-effector") %>% 
-    mutate(
-      data = data_origin, 
-      tool = tool_used
-    )
-  
-  return(conf_mat_df)
-}
-
-# Get all accuracy and f1_score
-metrics_list <- yardstick::metric_set(
-  yardstick::accuracy,
-  yardstick::f_meas
-)
-
-# Calculate accuracy and f1_score
-acc_f1_score_deept3_data_prediction <- rbind(
-rbind(
-  get_true_pred_value_from_deepredeff("data/raw/deept3_P.syringae_nr_effector_data_evaluated_on_deepredeff.csv", label_data = "effector") %>% 
[…]

### 2020-06-11 data/processed/compare_result_effectR_data_prediction.R

@@ -1,97 +1,0 @@
-library(dplyr)
-
-# Function to get true and pred from deepredeff output
-get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
-  df <- data.table::fread(data_path) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  return(df)
-}
-
-get_confusion_mat <- function(data_frame, filter){
-  
-  
-  conf_mat_df <- data_frame %>%
-    dplyr::filter(tool == filter) %>%
-    yardstick::conf_mat(truth = label, estimate = value) %>% 
-    .$table %>% 
-    as.data.frame() %>% 
-    filter(Truth != "non-effector") %>% 
-    mutate(
-      tool = filter
-    )
-  
-  return(conf_mat_df)
-}
-
-
-# Get all accuracy and f1_score
-metrics_list <- yardstick::metric_set(
-  yardstick::accuracy,
-  yardstick::f_meas
-)
-
-
-
-effectR_data_evaluated_on_deepredeff_and_effectR <- get_true_pred_value_from_deepredeff(
-  "data/raw/effectR_reference_rxlr_effectors_evaluated_on_deepredeff.csv", 
-  label_data = "effector"
-) %>% 
-  mutate(
-    name = sub(" .*", "", name)
-  ) %>% 
-  select(name, label, pred_cnn_lstm = pred) %>% 
-  left_join(
-    data.table::fread("data/raw/effectR_reference_rxlr_effectors_evaluated_on_effectR.csv") %>% 
-      select(name, pred_effectR = candidate), 
-    by = "name"
-  ) %>% 
-  dplyr::mutate(
-    pred_effectR = ifelse(is.na(pred_effectR), "non-effector", "effector")
-  ) %>% 
-  tidyr::pivot_longer(
-    cols = -c(name, label),
-    names_to = "tool"
-  ) %>% 
-  dplyr::mutate(
-    value = factor(value, levels = c("effector", "non-effector")),
-    label = factor(label, levels = c("effector", "non-effector"))
-  ) 
-
[…]

### 2020-06-11 data/processed/compare_result_effectivet3_data_prediction.R

@@ -1,97 +1,0 @@
-library(dplyr)
-
-# Function to get true and pred value from effectorp output
-get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
-  df <- data.table::fread(data_path) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  return(df)
-}
-
-
-get_true_pred_value_from_effectivet3 <- function(data_path, label_data){
-  
-  df <- data.table::fread(data_path, skip = 1) %>% 
-    dplyr::select(-c("Description", "Score")) %>% 
-    `colnames<-`(c("id", "is_effective")) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(is_effective == TRUE, 1, 0), levels = c(1, 0), labels = c("effector", "non-effector")), 
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  
-  return(df)
-}
-
-
-# Get confusion matrix from both
-get_confusion_mat <- function(data_path, label_data, data_origin, tool_used){
-  
-  if (tool_used != "ensemble_weighted"){
-    df <- get_true_pred_value_from_effectivet3(data_path, label_data = label_data)
-  } else {
-    df <- get_true_pred_value_from_deepredeff(data_path, label_data = label_data) 
-  }
-  
-  conf_mat_df <- df %>%
-    yardstick::conf_mat(truth = label, estimate = pred) %>% 
-    .$table %>% 
-    as.data.frame() %>% 
-    filter(Truth != "non-effector") %>% 
-    mutate(
-      data = data_origin, 
-      tool = tool_used
-    )
-  
-  return(conf_mat_df)
-}
-
-
-
-# Get all accuracy and f1_score
-metrics_list <- yardstick::metric_set(
-  yardstick::accuracy,
-  yardstick::f_meas
-)
-
-
-acc_f1_score_effectivet3_data_prediction <- rbind(
-rbind(
[…]

### 2020-06-11 data/processed/compare_result_effectorp_data_prediction.R

@@ -1,115 +1,0 @@
-library(dplyr)
-
-# Function to get true and pred value from effectorp output
-get_true_pred_value_from_effectorp_pred <- function(data_path, label_data){
-  df <- data.table::fread(data_path) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(Prediction == "Effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector")), 
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  return(df)
-}
-
-# Function to get true and pred from deepredeff output
-get_true_pred_value_from_deepredeff <- function(data_path, label_data) {
-  df <- data.table::fread(data_path) %>% 
-    dplyr::mutate(
-      pred = factor(ifelse(prob < 0.5, 0, 1), levels = c(1, 0), labels = c("effector", "non-effector")),
-      label = factor(ifelse(label_data == "effector", 1, 0), levels = c(1, 0), labels = c("effector", "non-effector"))
-    )
-  return(df)
-}
-
-# Get confusion matrix from both
-get_confusion_mat <- function(data_frame, label_data, data_origin, tool_used){
-  
-  if (tool_used != "cnn_lstm"){
-    df <- get_true_pred_value_from_effectorp_pred(data_frame, label_data = label_data)
-  } else {
-    df <- get_true_pred_value_from_deepredeff(data_frame, label_data = label_data) 
-  }
-  
-  conf_mat_df <- df %>%
-    yardstick::conf_mat(truth = label, estimate = pred) %>% 
-    .$table %>% 
-    as.data.frame() %>% 
-    filter(Truth != "non-effector") %>% 
-    mutate(
-      data = data_origin, 
-      tool = tool_used
-    )
-  
-  return(conf_mat_df)
-}
-
-# Get all accuracy and f1_score
-metrics_list <- yardstick::metric_set(
-  yardstick::accuracy,
-  yardstick::f_meas
-)
-
-
-# Calculate accuracy and f1_score
-acc_f1_score_effectorp_data_prediction <- rbind(
-  rbind(
-    get_true_pred_value_from_deepredeff("data/raw/effectorP1_pos_training_evaluated_on_deepredeff.csv", label_data = "effector") %>%
-      mutate(data = "effectorP1_training", tool = "cnn_lstm"), 
[…]

### 2020-06-11 functions/check_proportions_aas.R

@@ -27,13 +27,13 @@ get_global_counts <- function(df){
   
   counts <- get_counts(df)
   
-  list(
-    total_length = sum(counts$length),
-    total_hydrophobic = sum(counts$hydrophobic),
-    total_polar = sum(counts$polar),
-    total_neutral = sum(counts$neutral),
-    total_basic = sum(counts$basic),
-    total_acidic = sum(counts$acidic)
+  tibble::tibble(
+    all_length =  sum(counts$length),
+    all_hydrophobic = sum(counts$hydrophobic),
+    all_polar = sum(counts$polar),
+    all_neutral =sum(counts$neutral),
+    all_basic = sum(counts$basic),
+    all_acidic =  sum(counts$acidic)
   )
   
 }
@@ -42,13 +42,13 @@ get_effector_counts <- function(df, keep_label = 1){
   
   counts <- df %>% filter(label == keep_label) %>% 
     get_counts()
-  list(
-    total_length = sum(counts$length),
-    total_hydrophobic = sum(counts$hydrophobic),
-    total_polar = sum(counts$polar),
-    total_neutral = sum(counts$neutral),
-    total_basic = sum(counts$basic),
-    total_acidic = sum(counts$acidic)
+  tibble::tibble(
+    eff_length =  sum(counts$length),
+    eff_hydrophobic = sum(counts$hydrophobic),
+    eff_polar = sum(counts$polar),
+    eff_neutral =sum(counts$neutral),
+    eff_basic = sum(counts$basic),
+    eff_acidic =  sum(counts$acidic)
   )
 }
 
@@ -64,13 +64,76 @@ do_hyper <- function(df) {
 
   all_seqs <- get_global_counts(df)
   eff_seqs <- get_effector_counts(df)
-  tibble::tibble(
-    Hydrophobic = phyper(eff_seqs$total_hydrophobic, all_seqs$total_hydrophobic, (all_seqs$total_length - all_seqs$total_hydrophobic), eff_seqs$total_length ),
-    Polar = phyper(eff_seqs$total_polar, all_seqs$total_polar, (all_seqs$total_length - all_seqs$total_polar), eff_seqs$total_length ),
-    Neutral = phyper(eff_seqs$total_neutral, all_seqs$total_neutral, (all_seqs$total_length - all_seqs$total_neutral), eff_seqs$total_length ),
-    Basic = phyper(eff_seqs$total_basic, all_seqs$total_basic, (all_seqs$total_length - all_seqs$total_basic), eff_seqs$total_length ),
[…]

### 2020-06-12 main.Rmd
update method and results for hyperparameter scan

@@
 Phytopathogens are a major threat to global crop production.
-The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of ~6 billion USD to potato production, annually [@haas2009genome].
+The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of around 6 billion USD to potato production, annually [@haas2009genome].
 The bacterium _Ralstonia solanacearum_ has a wide host range and can cause loses of over 30% in potato, banana and groundnut [@Yuliar:2015hp].
@@
 We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM.
+^[Ruth, what about the stuff you did on the cluster with the GPUs?
+This isn't correct, is it?].
 
@@
 
-We performed a random search of hyperparameter effects by selecting 50 of all possible hyperparameters depending on each model type at random.
-The total of possible hyperparameter combination for each model can be found in Table \@ref(tab:possiblecombination).
-Additionally, the detail list of hyperparamaters that were fixed and tuned can be found in Table \@ref(tab:fixedhyperparam) and \@ref(tab:tunedhyperparam), respectively.
-We applied five-fold cross validation at each combination and recorded model performance on the validation data at each iteration.
-Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
-Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam).
+Hyperparameter scans were performed in ^[Ruth, can you outline the packages and methods in scikit-learn/tensorflow that you used for this] Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
+All modle training was performed as described in section Model Training.
 
@@
 
-With best performing models from hyperparameter scans we carried out a process of fine tuning manually to ensure that models did not over- or underfit data and were trained to be as general as possible.
-We loaded best performing models and assessed changes in selected parameters as described in Results to produce final models.
-These processes were performed manually using Python scripts that can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).
-The results were visualised using R and the codes are stored in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/r-scripts/model-scripts/multi-class/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/r-scripts/model-scripts/multi-class/secreted_data).
-Once each model was tuned accordingly using the training and test sets, final classification testing and estimation of accuracy was done with the held-out validation set that the model had not seen until this point.
-The final models were saved into HDF5 objects and stored in the repository at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5).
+Fine tuning was performed manually using Tensorflow/scikit learn ^[Ruth, can you add the software and versions your scripts used here] Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).
 
@@
 Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly.
-The total number of plant effectors per group can be seen in Table \@ref(tab:effsummary).
-The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable)
+In total 190 bacterial effectors from 13 species were collected, 97 fungal effectors from 16 species were collected and 85 oomycete effectors from 6 species were collected (Table \@ref(tab:effsummary) ).
+The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable).
 
@@
 Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool.
-For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected an identical number of non-effector, secreted proteins to that collected in the effector set from either the same strain or species.
+For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected from either the same strain or species an identical number of non-effector, secreted proteins to that in the effector set.
 This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples.
@@
 ## Model Selection and Training
-
+In order to identify a useful classifier we took a randomised hyperparamter search over some likely base model architectures.
 We selected four base architectures on which to build models for learning.
@@
 See Figure \@ref(fig:modelarch).
-Hyperparameter values for the models were identified using a random search as described in methods.
-Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
-For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.
-Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), a total of twelve models were produced, one for each taxon and model combination.
-The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).
+
+We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding.
+To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 120 * 50` models in total.
+^[Ruth, can you confirm that this interpretation is correct]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
+Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam) ^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction, the best performing values can be see in Table \@ref(tab:possiblecombination).
+From this we had twelve candidate models to develop further.
 
@@
 
-We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality.
+We then manually ran and checked the accuracy and loss of the twelve models on the training and validation sets to investigate instances of overfitting and assess generality.
 Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size.
+Parameters varied in this phase are listed in \@ref(tab:tunedhyperparam).
+^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam).
+The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).
 We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.
+
 Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
@@
 All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
+
+The final twelve models were saved into HDF5 objects and stored in the repository at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5).
 

### 2020-06-12 main.Rmd
add discussion

@@
 
-We have compiled a set of known experimentally validated effectors from phytopathogenic bacteria, fungi and oomycetes and used them as positive training examples with which to tune a range of CNN classifiers.
+We have compiled a set of known experimentally validated effectors from phytopathogenic bacteria, fungi and oomycetes and used them as positive training examples with which to tune a range of CNN classifiers, the sequences are all taken from the database PHI-Base, a manually curated database that aims to search all the literature on pathogen and host interactions [@phibase].
+The data in PHI-Base is complete as far as 2013 at the moment, thus the phytopathogen effectors we have collected should be all those shown experimentally to have an effect on a host that have been reported on since that time.
+We chose this set as we believed that this would give us the most reliable and unbiased set of effectors on which we could train learners.
+That is not to say that the set itself cannot be biased and that the set does not introduce any bias into the classifications of our learners.
+Sources of bias in our sequence set include the time limits on what has so far been included in PHI-Base, any effectors known but not reported on in this seven year time period cannot be represented in the models.
+The species of phytopathogens represented in the set also create bias, the effectors are not selected from species sampled proportionally or randomly but instead are those that trends in effector research over the last seven years have brought to focus.
+In particular, species that have had genome sequencing projects over this time are over-represented.
+There may also be some echoes of other methods previously applied, effectors studied experimentally must be identified first as hypothetical effectors, usually with the aid of computational tools whose models are themselves biased towards the sort of sequence that we already know.
+The effectors in the literature may therefore be enriched with respect to known features and results from classifiers should be interpreted with this in mind.
+We are not easily able to quantify any bias, but the greater generalisability of our models over the others tested gives us reason to believe that the sequence set trained on was broad enough for good models to be developed.
+
+A common misconception of deep learning models is that training datasets need to be extremely large, this is only half true.
+In fact, in practice training data need only to be large with respect to the model size.
+Here we have coupled small data sets (tens or hundreds of training examples) with small models.
+The layers and architectures of the models presented here are much smaller than those typically used in large machine vision projects for example.
+Yet the small models remain useful and have predictive power, indicating a definite role for deep learning approaches in the work of bioinformaticians without truly massive datasets.
+
+Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[Ruth, have we any idea what this fraction might be] of the possible hyperparameter and parameter space was explored we still compared 6000 models.
+At a run time of around ^[Ruth, approx run time per model] So even for relatively small data sets, a large amount of specialised GPU compute power is required.
+
+The models we created performed exceptionally on the PHI-Base hold-out validation data set of phytopathogenic effector sequences.
+The greater than 80 % classification accuracy for the fungal and oomycete models is an excellent accuracy on such sequences and our models outperformed the other classifiers tested by large margins on the PHI-Base fungal, no other machine learning method has been reported to have performed as well on phytopathogen sequences.
+The greater than 97 % accuracy we observed in our model trained on bacterial phytopathogen effectors is also exceptional and similar to what Xue et al [@deept3] showed in human bacterial pathogen effectors.
+When we evaluated the proportion of effectors each of our and other classifiers could find in data used to create the other classifiers, we found that our bacterial model outperformed the bacterial models again.
+A slightly different picture emerged when we compared our method with EffectorP 1 and 2 on fungal data.
+Both versions of Effector P outperformed our model on training data used to create Effector P in the first instance but identical predictions were made by all fungal classifiers on the validation set provided by Effector P 2.
+This was coincident with a large drop in accuracy from training to validation data by Effector P 1 and 2.
+Combined with observations on our PHI-Base data on higher numbers of false positives from Effector P, we conclude that Effector P is over-fitted slightly on its training data and that our model is likely to be more generally accurate.
+We expect that the tools used in concert will provide very good classification and prediction of fungal effectors.
+Our model also performed worse on the data used to test the untrained heuristic oomycete RXLR sequence detector EffectR, finding three of six, relative to EffectR's five of six.
+This is likely due to the six test sequences being exclusively RXLR and our model being trained on effectors in general, the presence of RXLR not being diagnostic of effector-ness in our model means Researchers hoping to find exclusively RXLR containing proteins would be advised to use EffectR, those hoping to find effectors beyond RXLRs may find utility with our oomycete model.
+
+In developing a negative training data set that contained secreted non-effector proteins we hoped to decouple the identification of effector from identification of secreted protein as a proxy for effector.
+We believe that we have been succesful at this, the models we developed do identify effectors against a background of secreted proteins, indicating that they have some internal representation of what makes an effector different from a secreted peptide.
+By examining the activation maps of the CNN models, we learned that the maximum activations in the models remains in the N-terminus of the proteins, coincident with the expected positions of secretions signals and is relatively narrow (about 25 amino acids.
+We also noted that there is no typical primary sequence motif that can be identified, the identity of the amino acids themselves does not seem important.
+We did find that various categories of amino acid were enriched or depleted significantly in the narrow activation region of effectors, relative to the same area of non-effector secreted proteins.
+A favouring of functional properties of the amino acids of the effector over the actual identity may be reflective of the many functional types origins of effectors.
+It may be that the N-terminal activation region in the effectors represents non-canonical secretion-compatible signal in the effector proteins.
+In order to evolve towards a useful secretion signal from an arbitrary point it may be enough to have a region that satisfies some chemical property that makes it secretable, yet is different enough in effectors to separate them from the secreted proteins by a deep learning algorithm.
+
+Finally, we have made our models available in an R package _deepredeff_ - from 'deep learning prediction of effector'.
+This can be obtained from GitHub as shown in methods, installation and usage instructions are available in the documentation provided there.
+The R package allows the user to run the sequences from various sequence format files against a selected model and obtain a _p_-value that each sequence is an effector according to the model.
+Various summaries and plots are provided for the user when working on large numbers of sequences.
+The package integrates well with Bioconductor.
+For those wishing to use Python to make use of our models, we provide the models as HDF5 files from TensorFlow that can be use in that toolkit.
 

### 2020-06-15 main.Rmd
added some missing parts

@@
 We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding.
-We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM.
-^[Ruth, what about the stuff you did on the cluster with the GPUs?
-This isn't correct, is it?].
+We trained each model using a basic random hyperparameter setting initialisation step followed by hyperparamter scans.
+All models were implemented in Python 3.6.9 using the deep learning API Keras 2.2.4 with Tensorflow 1.12.0 backend, using GPU NVIDIA GK110GL Tesla K20c and CPU AMD Opteron(TM) Processor 6272 with memory 128 GB.
 
@@
 
-Hyperparameter scans were performed in ^[Ruth, can you outline the packages and methods in scikit-learn/tensorflow that you used for this] Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
-All modle training was performed as described in section Model Training.
+Hyperparameter scans were performed using random search, a hyperparameter optimization method which 
+implements a randomized search over hyperparameters, where each hyperparameter setting is randomly sampled from a distribution over possible hyperparameter values [@Bergstra2012].
+In particularly, we used RandomSearchCV(), the implementation of random search method in `scikit-learn 0.19.2` [@scikit-learn] together with KerasClassifier() which is implementation of the scikit-learn classifier API for keras.
+Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
+All model training was performed as described in section Model Training.
 
@@
 
-Fine tuning was performed manually using Tensorflow/scikit learn ^[Ruth, can you add the software and versions your scripts used here] Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).
+Fine tuning was performed manually using `keras 2.2.4` together with metrics module and KFold cross validator from `scikit-learn 0.19.2`.
+Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).
 
@@
 To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we adapted a Grad-CAM (Gradient-weighted Class Activation Mapping) approach [@Selvaraju2019].
-We extracted the convolutional layer `convd_1` and computed the gradient of the model output with respect to the output of `convd_1` The feature map was weighted by every channel and mean result taken.
-All activation maps were summed, normalised and smoothed using Fourier analysis ^[Ruth, there need to be references, mentions of software and versions used, it isn't easy to see what you actually did from this description].
+We extracted the convolutional layer `convd_1` and computed the gradient of the model output with respect to the output of `convd_1`.
+The feature map was weighted by every channel and mean result taken.
+All activation maps were summed, normalised and smoothed using Fourier analysis.
+We used `tensorflow 1.12.0` to compute the activation maps and discrete fourier transform (numpy.fft) from `numpy 1.17.3` to smooth the result.
 The code we used for computing and visualising these heatmaps can be found at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .
@@
 We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding.
-To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 120 * 50` models in total.
-^[Ruth, can you confirm that this interpretation is correct]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
-Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam) ^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction, the best performing values can be see in Table \@ref(tab:possiblecombination).
-From this we had twelve candidate models to develop further.
+To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 3 * 4 * 50 * 5` models in total.
+^[For each architecture, 50 models are selected randomly.
+Therefore there are 3 taxon * 4 architecture * 50 models * 5 cross validation]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
+Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam).
+For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction.
+^[\@ref(tab:possiblecombination) is table of total number of all hyperparameter settings in the whole hyperparameter space for each architecture] From this we had twelve candidate models to develop further.
 
@@
 Parameters varied in this phase are listed in \@ref(tab:tunedhyperparam).
-^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam).
+Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam).
 The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).
@@
 
-Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[Ruth, have we any idea what this fraction might be] of the possible hyperparameter and parameter space was explored we still compared 6000 models.
-At a run time of around ^[Ruth, approx run time per model] So even for relatively small data sets, a large amount of specialised GPU compute power is required.
+Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[This fraction depends for every model, the parameter space is shown in table \@ref(tab:possiblecombination)] of the possible hyperparameter and parameter space was explored we still compared `r 3 * 4 * 50 * 5` models.
+At a run time of around 144 minutes for CNN-LSTM, 57.2 minutes for CNN-GRU, and 45 minutes for both GRU Embedding and LSTM Embedding.
+So even for relatively small data sets, a large amount of specialised GPU compute power is required.
 
@@
 ```{r, fixedhyperparam}
-apa_table(
-  readr::read_tsv("tables/fixed_hyperparam.tab"),
+# apa_table(
+#   readr::read_tsv("tables/fixed_hyperparam.tab"),
+#   digits = 0,
+#   align = 'c',
+#   caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
+# )
+
+readr::read_tsv("tables/fixed_hyperparam.tab") %>%
+  knitr::kable(
+    "latex", 
+    longtable = T, 
+    booktabs = T,
     digits = 0,
@@
     caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
-)
+  ) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
 ```
@@
 ```{r, tunedhyperparam}
-apa_table(
-  readr::read_tsv("tables/tuned_hyperparam.tab"),
+# apa_table(
+#   readr::read_tsv("tables/tuned_hyperparam.tab"),
+#   digits = 0,
+#   align = 'c',
+#   caption = 'Hyperparameters tuned in manual scanning of hyperparameter space for each model'
+# )
+
+readr::read_tsv("tables/tuned_hyperparam.tab") %>%
+  knitr::kable(
+    "latex", 
+    longtable = T, 
+    booktabs = T,
     digits = 0,
@@
     caption = 'Hyperparameters tuned in manual scanning of hyperparameter space for each model'
-)
+  ) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
 ```
@@
 ```
-
-```

### 2020-06-15 main.Rmd
add abstract

@@
   - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
-abstract: "TBD."
+abstract: "Plant pathogens cause billions of dollars of crop loss every year and are a major threat to global food security.
+Effector proteins are the tools such pathogens use to infect the cell, predicting effectors de novo from sequence is difficult because of the heterogeneity of the sequences.
+We hypothesised that deep learning classifiers based on Convolutional Neural Networks would be able to identify effectors and deliver new insights.
+We built a training set of manually curated effector sequences from PHI-Base and used these to train a range of model architectures for classifying bacteria, fungal and oomycete sequences.
+The best performing classifiers had accuracies from 93 % to 84 %.
+The models were tested against popular effector detection software on our own test data and data provided with those models.
+We observed better performance from our models.
+Specifically our models showed greater accuracy and lower tendencies to call false positives on a secreted protein negative test set and a greater generalisability.
+We used GRAD-CAM activation map analysis to identify the sequences that activated our CNN-LSTM models and found short but distinct N-terminal regions in each taxon that was indicative of effector sequences.
+No motifs could be observed in these regions but an analysis of amino acid types indicated differing patterns of enrichment and depletion that varied between taxa.
+We have produced an R package that will allow others to make easy effector predictions using our models."
 geometry: margin=1in
@@
 We trained each model using a basic random hyperparameter setting initialisation step followed by hyperparamter scans.
-All models were implemented in Python 3.6.9 using the deep learning API Keras 2.2.4 with Tensorflow 1.12.0 backend, using GPU NVIDIA GK110GL Tesla K20c and CPU AMD Opteron(TM) Processor 6272 with memory 128 GB.
+All models were implemented in Python 3.6.9 [@python] using the deep learning API Keras 2.2.4 [@chollet2015] with the Tensorflow 1.12.0 [@tensorflow2015] backend, using  NVIDIA GK110GL Tesla K20c GPUs and AMD Opteron(TM) Processor 6272 CPUs with 128 GB RAM.
 
@@
 
-Hyperparameter scans were performed using random search, a hyperparameter optimization method which 
-implements a randomized search over hyperparameters, where each hyperparameter setting is randomly sampled from a distribution over possible hyperparameter values [@Bergstra2012].
-In particularly, we used RandomSearchCV(), the implementation of random search method in `scikit-learn 0.19.2` [@scikit-learn] together with KerasClassifier() which is implementation of the scikit-learn classifier API for keras.
+Hyperparameter scans were performed using random search, a hyperparameter optimization method where each hyperparameter setting is randomly sampled from a distribution of possible hyperparameter values [@Bergstra2012].
+We used RandomSearchCV(), the implementation of random search in `scikit-learn 0.19.2` [@scikitlearn] together with KerasClassifier() which is an implementation of the scikit-learn classifier API for keras.
 Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts).
@@
 To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 3 * 4 * 50 * 5` models in total.
-^[For each architecture, 50 models are selected randomly.
-Therefore there are 3 taxon * 4 architecture * 50 models * 5 cross validation]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
+Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data.
 Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam).
 For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction.
-^[\@ref(tab:possiblecombination) is table of total number of all hyperparameter settings in the whole hyperparameter space for each architecture] From this we had twelve candidate models to develop further.
+From this we had twelve candidate models to develop further.
 
@@
 
-Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[This fraction depends for every model, the parameter space is shown in table \@ref(tab:possiblecombination)] of the possible hyperparameter and parameter space was explored we still compared `r 3 * 4 * 50 * 5` models.
-At a run time of around 144 minutes for CNN-LSTM, 57.2 minutes for CNN-GRU, and 45 minutes for both GRU Embedding and LSTM Embedding.
-So even for relatively small data sets, a large amount of specialised GPU compute power is required.
+Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised (see Table \@ref(tab:possiblecombination) for a breakdown) and although only a small fraction  of the possible hyperparameter and parameter space was explored we compared `r 3 * 4 * 50 * 5` models, at a run time of around 144 minutes for CNN-LSTM, 57.2 minutes for CNN-GRU, and 45 minutes for both GRU Embedding and LSTM Embedding.
+For these relatively small data sets, a significant amount of specialised GPU compute power was required.
 

### 2020-06-17 main.Rmd
typos

@@
 Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
-We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel).
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we observed accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel).
 All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
@@
 CNN-GRU and LSTM-Embedding showed identical prediction sets.
-We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51.
+We observed similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51.
 The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65.
@@
 We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults).
-With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
-However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures.
+With the models trained in bacterial sequences we observed an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
+However, with the fungal and oomycete models we observed decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures.
 Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models.
@@
 Each comparison was carried out using the respective hold-out test sequence set for each taxon.
-For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics).
+For all taxa we observed greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics).
 This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
@@
     aes(Tool, `Proportion Found`) +
-    geom_bar(stat= 'identity', fill = rgb(52/255, 106/255, 141/255), colour = rgb(52/255, 106/255, 141/255), alpha=0.5) +
+    geom_bar(stat= 'identity', fill = rgb(52/255, 106/255, 141/255), colour = rgb(52/255, 106/255, 141/255)) +
     facet_grid(  ~ `Data Set`, scales = "free") + 

### 2021-03-24 main.Rmd
edits for BMC submission

@@
   - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
-abstract: "Plant pathogens cause billions of dollars of crop loss every year and are a major threat to global food security.
-Effector proteins are the tools such pathogens use to infect the cell, predicting effectors de novo from sequence is difficult because of the heterogeneity of the sequences.
-We hypothesised that deep learning classifiers based on Convolutional Neural Networks would be able to identify effectors and deliver new insights.
-We built a training set of manually curated effector sequences from PHI-Base and used these to train a range of model architectures for classifying bacteria, fungal and oomycete sequences.
-The best performing classifiers had accuracies from 93 % to 84 %.
-The models were tested against popular effector detection software on our own test data and data provided with those models.
-We observed better performance from our models.
-Specifically our models showed greater accuracy and lower tendencies to call false positives on a secreted protein negative test set and a greater generalisability.
-We used GRAD-CAM activation map analysis to identify the sequences that activated our CNN-LSTM models and found short but distinct N-terminal regions in each taxon that was indicative of effector sequences.
-No motifs could be observed in these regions but an analysis of amino acid types indicated differing patterns of enrichment and depletion that varied between taxa.
-We have produced an R package that will allow others to make easy effector predictions using our models."
+abstract: ""
 geometry: margin=1in
@@
 theme_set(theme_bw())
+
+fig_caption <- captioner::captioner(prefix = "Figure ")
+fig_caption("effectorcollection", "Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt")
+
+fig_caption("noneffectorcollection", "Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists")
+
+fig_caption("modelarch", "Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model")
+
+fig_caption("trainingprogress", "Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set")
+
+fig_caption("overallmodel", "Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.")
+
+fig_caption("confusionmodelcorrs", "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models.
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.")
+
+fig_caption("ensembleresults", "Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.")
+
+fig_caption("vsothersmetrics", "Accuracy and F1 score of the classifications made by our models compared with those made by widely used software")
+
+fig_caption("vsotherscfmat", "Confusion matrices of classifications made by our models compared with those made by widely used software.
+A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.")
+
+fig_caption("owndata", "Performance of models on training and validation datasets used to train other effector prediction tools.
+A: Tools trained on bacterial sequences; B: Tools trained on fungal sequences; C: Tools trained on oomycete sequences.")
+
+fig_caption("activations", "Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data (left panel) and non-effector data (right panel).
+A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.")
+
+tbl_caption <- captioner::captioner(prefix = "Table ")
+tbl_caption("effsummary", 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups')
+
+tbl_caption("aclossfinal", 'Accuracy and loss values from best performing parameters values for each model')
+
+tbl_caption("phyperaatype", 'Enrichment analysis of amino acid types.
+Proportions of each amino acid type in the activation region of All Sequences : Effector Sequences, with (Probability).
+Probability by hypergeometric test of the observed number of amino acid type in the region of activation in effector sequences in our compiled sequence set relative to a background of the same region in all sequences in that taxon.')
 ```
@@
 ```
+\newpage
+# Abstract
 
-# Introduction
+## Background:
+
+Plant pathogens cause billions of dollars of crop loss every year and are a major threat to global food security.
+Effector proteins are the tools such pathogens use to infect the cell, predicting effectors de novo from sequence is difficult because of the heterogeneity of the sequences.
+We hypothesised that deep learning classifiers based on Convolutional Neural Networks would be able to identify effectors and deliver new insights.
+
+## Results:
+
+We created a training set of manually curated effector sequences from PHI-Base and used these to train a range of model architectures for classifying bacteria, fungal and oomycete sequences.
+The best performing classifiers had accuracies from 93 % to 84 %.
+The models were tested against popular effector detection software on our own test data and data provided with those models.
+We observed better performance from our models.
+Specifically our models showed greater accuracy and lower tendencies to call false positives on a secreted protein negative test set and a greater generalisability.
+We used GRAD-CAM activation map analysis to identify the sequences that activated our CNN-LSTM models and found short but distinct N-terminal regions in each taxon that was indicative of effector sequences.
+No motifs could be observed in these regions but an analysis of amino acid types indicated differing patterns of enrichment and depletion that varied between taxa.
+
+## Conclusions:
+
+Small training sets can be used effectively to train highly accurate and sensitive deep learning models without need for the operator to know anything other than sequence and without arbitrary decisions made about what sequence features or physico-chemical properties are important.
+Biological insight on subsequences important for classification can be achieved by examining the activations in the model
+
+\newpage
+# Background
 
@@
 
-# Methods
+# Implementation
 
@@
 Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
-The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used.
+The pipeline in Figure \@ref(effectorcollection) outlines the steps used.
 We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd).
@@
 
+## Software Implementation
+
+To make our models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
+The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff) and CRAN [https://cran.r-project.org/package=deepredeff](https://cran.r-project.org/package=deepredeff)
+
 # Results
@@
 This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifically reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal.
-To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID.
+To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID (see `r fig_caption("effectorcollection", display="cite")`.
 The sequences and IDs retrieved can be seen in the data file in this manuscript's repository [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv).
@@
 This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples.
-Figure \@ref(fig:noneffectorcollection) summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).
+`r fig_caption("noneffectorcollection", display="cite")` summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).
 
-```{r effectorcollection, fig.width=10, fig.height=10, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
+```{r effectorcollection, fig.width=10, fig.height=10, out.width='50%', fig.align='center', fig.cap=fig_caption("effectorcollection",display="cite")}
 knitr::include_graphics(
@@
 
-```{r noneffectorcollection, fig.width=10, fig.height=10, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
+```{r noneffectorcollection, fig.width=10, fig.height=10, fig.cap=fig_caption("noneffectorcollection",display="cite")}
 knitr::include_graphics(
@@
   align = 'c',
-  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
+  caption = " "
 )
@@
 All models had fully-connected dense layers after this.
-See Figure \@ref(fig:modelarch).
+See  `r fig_caption("modelarch", display="cite")`.
 
@@
 
-```{r modelarch, fig.width=10, fig.height=10, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C: LSTM-Embedding model; D: GRU-Embedding model"}
+```{r modelarch, fig.width=10, fig.height=10, fig.cap=fig_caption("modelarch",display="cite")}
 knitr::include_graphics("figures/model_arch.pdf")
@@
   align = 'c',
-  caption = 'Accuracy and loss values from best performing parameters values for each model'
+  caption = ' '
          )
@@
 
-Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress).
-We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we observed accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel).
+Final training progressions for each model in each taxon can be seen in `r fig_caption("trainingprogress", display="cite")`.
+We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we observed accuracy from 64.7 to 82.3, reported in `r fig_caption("overallmodel", display="cite")`.
 All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.
@@
 
-```{r trainingprogress, fig.height= 10, fig.width= 10, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
+```{r trainingprogress, fig.height= 10, fig.width= 10, fig.cap=fig_caption("trainingprogress",display="cite")}
 # knitr::include_graphics("figures/final_train_progress.png")
@@
 
-```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r overallmodel, fig.cap=fig_caption("overallmodel",display="cite")}
 # apa_table(
@@
 
-Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs).
+Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see  `r fig_caption("confusionmodelcorrs", display="cite")`.
 CNN-GRU and LSTM-Embedding showed identical prediction sets.
@@
 
-```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models.
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
+```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = fig_caption("confusionmodelcorrs",display="cite"), message = FALSE, echo=FALSE}
 
@@
 We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input.
-We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults).
+We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in `r fig_caption("ensembleresults", display="cite")`.
 With the models trained in bacterial sequences we observed an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions.
@@
 
-```{r ensembleresults, fig.width=8, fig.height=4, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
+```{r ensembleresults, fig.width=8, fig.height=4, fig.cap=fig_caption("ensembleresults",display="cite")}
 # apa_table(
@@
 Each comparison was carried out using the respective hold-out test sequence set for each taxon.
-For all taxa we observed greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics).
+For all taxa we observed greater Accuracy and F1 scores from our models than the established software, as shown in `r fig_caption("vsothersmetrics", display="cite")`.
 This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls.
 Absolute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences.
-The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
+The confusion matrices in  `r fig_caption("vsotherscfmat", display="cite")` show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.
 All the established software classifiers we tested show lower sensitivity than the models we have developed here.
@@
 We calculated the proportion of the effectors in the training set that the tools could find on their respective training and validation sets, according to availability.
-The bacterial tools EffectiveT3 and DeepT3 showed lower proportion found than our Ensemble Weighted model, as in Figure \@ref(fig:owndata)A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated.
-Interestingly, both versions of EffectorP found a greater proportion of the effectors in the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically (Figure \@ref(fig:owndata)B ).
+The bacterial tools EffectiveT3 and DeepT3 showed lower proportion found than our Ensemble Weighted model, as in `r fig_caption("owndata", display="cite")`A, consistent with the observation that our Ensemble Weighted model performed more strongly on the validation set that we generated.
+Interestingly, both versions of EffectorP found a greater proportion of the effectors in the EffectorP provided training sets than our CNN-LSTM model, but in the unseen validation data provided with EffectorP 2, all three models performed identically ( `r fig_caption("owndata", display="cite")`B ).
 The EffectorP 1 and 2 scores on validation data are well below the scores for the training data, a result that is usually interpreted as being evidence of an overfitted and less generalisable model.
@@
 As this is not a trainable model in the same sense as the others, no large training set is needed.
-We attempted classification with these and effectR was able to classify 5 of the 6, whereas our CNN-LSTM model for oomycetes classified 3 of the 6 ( Figure \@ref(fig:owndata)C ).
+We attempted classification with these and effectR was able to classify 5 of the 6, whereas our CNN-LSTM model for oomycetes classified 3 of the 6 (  `r fig_caption("owndata", display="cite")`C ).
 
@@
 
-```{r vsothersmetrics, fig.width=8, fig.height=6, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", fig.pos="H"}
+```{r vsothersmetrics, fig.width=8, fig.height=6, fig.cap=fig_caption("vsothersmetrics",display="cite"), fig.pos="H"}
 data.table::fread("data/processed/acc_f1score_vs_othertools.csv") %>%
@@
 
-```{r vsotherscfmat, fig.width=8, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software.
-A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", fig.pos="H"}
+```{r vsotherscfmat, fig.width=8, fig.height=8, fig.cap=fig_caption("vsotherscfmat",display="cite"), fig.pos="H"}
 
@@
 
-```{r, owndata, fig.height = 8, fig.cap="Performance of models on training and validation datasets used to train other effector prediction tools.
-A: Tools trained on bacterial sequences; B: Tools trained on fungal sequences; C: Tools trained on oomycete sequences."}
+```{r, owndata, fig.height = 8, fig.cap=fig_caption("owndata",display="cite")}
 d <- readr::read_csv(here::here("data","processed", "owndatasets.csv"))
@@
 To examine the responses of the models, we ran all sequence data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method.
-The profiles were smoothed using FFT and examined, (see Figure \@ref(fig:activations) ).
+The profiles were smoothed using FFT and examined, (see `r fig_caption("activations", display="cite")` ).
 All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals.
@@
 We examined further the sequences under the largest peaks, specifically for bacterial sequences we used amino acids 25 to 50, for fungal sequences we used amino acids 35 to 80 and for oomycete sequences we used amino acids 15 to 40.
-Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs as shown in the logo plots in Fig \@ref(fig:activations) even within a taxon and data set.
+Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs as shown in the logo plots in `r fig_caption("activations", display="cite")` even within a taxon and data set.
 Enrichment analysis of different amino acid categories showed some statistically significant ($p < 0.05$) changes in proportions of amino acid types relative to the whole set of sequences in the activated regions (Table \@ref(tab:phyperaatype) ).
@@
 
-```{r activations, fig.height=10, fig.width=10, fig.cap="Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data (left panel) and non-effector data (right panel).
-A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
+```{r activations, fig.height=10, fig.width=10, fig.cap=fig_caption("activations",display="cite")}
 # knitr::include_graphics("figures/convolution_maps_fixed.pdf")
@@
   font_size = 'footnotesize',
-  caption = 'Enrichment analysis of amino acid types.
-Proportions of each amino acid type in the activation region of All Sequences : Effector Sequences, with (Probability).
-Probability by hypergeometric test of the observed number of amino acid type in the region of activation in effector sequences in our compiled sequence set relative to a background of the same region in all sequences in that taxon.'
+  caption = ' '
 )
@@
 ```
-
-## Software Implementation
-
-To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the models.
-The package and installation instructions are available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)
-
-<!--
-
-#########################################################
-## Start code fragments
-##
-##
-#########################################################
-
--->
 
@@
 For those wishing to use Python to make use of our models, we provide the models as HDF5 files from TensorFlow that can be use in that toolkit.
+
+# List Of Abbreviations
+
+ML: Machine Learning,
+CNN: Convolutional Neural Network,
+MPMI: Molecular Plant Microbe Interactions,
+LSTM: Long Short Term Memory,
+GRU: Gated Recurrent Unit,
+CPU: Central Processing Unit,
+GPU: Graphics Processing Unit,
+TM: Trademark,
+TP: True Positive,
+FP: False Positive,
+FN: False Negative,
+TN: True Negative,
+Grad-CAM: Gradient-weighted Class Activation Mapping,
+ID: Identifier,
+FFT: Fast Fourier Transform,
+
+# Declarations
+
+## Availability of data and material
+
+The datasets generated and/or analysed during the current study are available in the 'ruth-effectors-prediction' repository, https://github.com/TeamMacLean/ruth-effectors-prediction.
+Individual datasets location within this repository are listed per dataset in the Results section
+
+## Competing Interests
+
+We have no competing interests.
+
+## Funding
+
+RKR and DM were supported by The Gatsby Charitable Foundation core grant to The Sainsbury Laboratory.
+
+## Authors' Contributions
+
+RKR carried out experiments and developed analysis code, and developed the software tools.
+DM designed experiments, interpreted results, and wrote the manuscript.
+All authors have read and approved the manuscript.
+
+## Acknowledgements
+
+We wish to thank Martin Page, Ram Krishna Shrestha, the NBI Computing Infrastucture for Science Team for computational support and Karen Smith for researcher support.
+We also to wish the staff of PHI-Base, particularly Dr Martin Urban for assistance in querying and bulkdownload from their database.
+
+\newpage
+
+# Figure Legends
+`r fig_caption("effectorcollection", display="full")`
+
+`r fig_caption("noneffectorcollection", display="full")`
+
+`r fig_caption("modelarch", display="full")`
+
+`r fig_caption("trainingprogress", display="full")`
+
+`r fig_caption("overallmodel",display="full")`
+
+`r fig_caption("confusionmodelcorrs",display="full")`
+
+`r fig_caption("ensembleresults",display="full")`
+
+`r fig_caption("vsothersmetrics",display="full")`
+
+`r fig_caption("vsotherscfmat",display="full")`
+
+`r fig_caption("owndata",display="full")`
+
+`r fig_caption("activations",display="full")`
+
+# Table Legends
+`r tbl_caption("effsummary", display="full")`
+
+`r tbl_caption("aclossfinal", display="full")`
+
+`r tbl_caption("phyperaatype",display="full")`
 
@@
 ```
+
+\newpage

### 2021-06-21 response_to_reviewers.Rmd
updates for reviewers

---
title: "Response To Reviewers Comments"
author: "Dan MacLean"
date: "21/06/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor,

Thank you for taking the time to send our manuscript for review. Please find attached a list of revisions in response to the reviewers comments that you provided.

We hope that you will find these satisfactory

Sincerely

Prof. Dan MacLean

## Reviewer 1

    1. 'The language in the manuscript needs close attention. For example, the sentence "Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as an effector or not.'

The manuscript was written by a native English speaker and we believe in general that the text is clear, as the same reviewer points out `the authors explained the concepts very clearly`. We have modified the sentence mentioned to read `Deep learning approaches require positive and negative examples from which to learn. We used a list of sequences annotated as an effector or not`

    2. 'This sentence "The larger and more accurate the list the more sensitivity a model can obtain." requires a citation of other works that explore the change in sensitivity with increasing training examples.

No such citation can be made as no study exists. This is because statement is one of a trivial fact about training set size. It is considered self-evident in the field that the more training examples you have, the better predictions can become. We have rephrased to `It is generally held and expected that he larger and more accurate the list the more sensitivity a model can obtain.`

    3. 'It will be nice to conduct a Wilcoxin's [sic] test to ensure that the difference in performance of the different classifiers are significant.'

We understand the impulse behind the suggestion but believe that this is not appropriate for the data presented. The per data set classifications can only be tested once, as they result in the same output for the same input every time, there is no experimental variability. Furthermore, it does not seem really informative to run hypothesis tests on these data, if the reviewer is suggesting pooling the accuracies we'd be testing the hypotheses that accuracy over all test data sets we used the tools on is equal. If one was significant over all data sets it would seem to suggest that the tool was 'better' overall. Our tool would likely come out better over all the datasets, but such a result would mask the more important results about relative performance across kingdoms and different data sets that could guide a user better. The confusion matrices are the best indicators of relative performance characteristics for a given test data set and we use those. Using a single hypothesis test across all data would leave us open to the justified criticism that we had cherry picked a single summary. This is not our intention, so we suggest not to perform this summary statistic and rely on the more explicit, detailed statistics that we have.

    4. 'It will be good to mention the location of the R package next to the sentence "We have produced an R package that will allow other scientists to easily classify their own sequences of interest."'

We have added the link to the R package homepage at the point the reviewer suggests, in addition to its existing places in the manuscript.

    5.  'Can this software be used to predict if a protein is an effector or not? You mentioned how the data was curated for each of the kingdoms. I wonder how the negative data would be constructed if you were to set it up to simply predict if a protein was an effector or not.'

For the first part, yes, this software can be used to predict if a protein is an effector. Our entire manuscript, including the title, and this reviewer point this out in their opening statement `In this manuscript the authors have explored a novel technique to predict effectors` We believe that this point is covered well enough if the reviewer did in fact notice it.

For the second part, we are not sure how to proceed. This seems to be a thought from the reviewer that we have already covered. The reviewer seems to be wondering how we might set up negative data in the case that we want to predict if a protein was an effector. As we do want to predict whether proteins are effectors, and did need negative data, we set it up as fully described in materials and methods, which the reviewer has noted we did. We do not think anything further needs doing for this point.

## Reviewer 2

    1. 'It is a common and most essential practice to train and test methods using different datasets. Authors used 60%, 20% and 20%  for train,test and validation. Is the selection random? How many times authors repeated it with different combination of train,test and validation tests.'

The selection was random at each model, we have updated the section `Encoding and Subsetting Sequences` to reflect this  now reading `Encoded sequences were split into random taxon specific training, test and validation`. The process over hyperparameter scans was repeated 3000 times, resulting in 3000 models being tested. This is described in section `Model Selection and Training`

    2. 'Did the authors checked the performance of the proposed method against the datasets used in the existing methods?'

Yes, we did. This is described in the dedicated section `Comparison With Other Classification Software` and highlighted as `We also evaluated the deep learning models we have developed on the training and hold-out validation sequences used to train the previous methods.` in the second paragraph. Figure 10 shows the results of this.

     3. 'Including a table that shows the sensitivity, specificity and accuracy of the proposed method and existing methods on test/validation dataset will be needed.'

The data are already described in confusion matrices in Figures 6, 9 and 10 and in Figure 8 along with the F1 score.

## Reviewer 3

    1. 'Did authors have performed 3 fold or 5 fold cross-validation?'

We did perform five fold cross-validation in the model training, we have added `Five-fold cross validation was used in all instances.` to the methods section.

    2. 'Authors have selected multiple hyperparameters for each classifiers, and mentioned that they have further chosen some of the hyperparameters to take multiple values and fixed rest of the hyperparameters for each classifier. What criteria were in mind or how authors have decided that which hyperparameters should be fixed and which not? This should be elaborated in methods section. GRID search - explain that.'

As described in the methods section and section `Model Selection and Training` (paragraph 2 and 3), we selected hyperparameters likely to have a strong effect and for those selected random values within likely useful ranges, running a model for each combination, a so-called random search. We present the range and values that we used in the associated supplemental tables. We are not sure why the reviewer has suggested we need to explain a grid search as it is not something we performed. We performed a random search of the space which we have described.

    3. 'Line 35 on page 4,missing figure number.'

This is corrected.

    4. 'Supplementary tables are not in correct order, please correct them.'

This is corrected.

    5. 'In figure 11, it is not clear  what does left, middle and right panel for each taxon represent. Make it more descriptive in figure caption'

We have updated the caption to read: `Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets showing the region of the sequence the model is most able to use to classify (leftmost panel). The most active part of the sequences are shown in logo plots of sequences extracted from the activation region within the dotted lines for effector data (centre panel) and non-effector data (rightmost panel). A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.`

### 2021-06-21 main.Rmd
updates for reviewers

@@
   - name: Ruth Kristianingsih, Dan MacLean
-  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
+  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7UH, dan.maclean@tsl.ac.uk
 abstract: ""
@@
   - \usepackage{float}
+  - \usepackage{placeins}
 ---
@@
 
-fig_caption("activations", "Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets; and sequence logo plots of sequence in the activation region within the dotted lines for effector data (left panel) and non-effector data (right panel).
+fig_caption("activations", "Grad-CAM activation profiles of CNN-LSTM models  when run with all datasets showing the region of the sequence the model is most able to use to classify (leftmost panel).
+The most active part of the sequences are shown in logo plots of sequences extracted from the activation region within the dotted lines for effector data (centre panel) and non-effector data (rightmost panel).
 A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.")
@@
 
-Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as an effector or not.
-The larger and more accurate the list the more sensitivity a model can obtain.
+Deep learning approaches require positive and negative examples from which to learn.
+We used a list of sequences annotated as an effector or not.
+It is generally held and expected that he larger and more accurate the list the more sensitivity a model can obtain.
 It is critical that training examples are experimentally verified effectors.
@@
 We also analyse the activations of the models in response to effectors and non-effectors to gain insights into the sequence features that are allowing classification.
-We have produced an R package that will allow other scientists to easily classify their own sequences of interest.
+We have produced an R package that will allow other scientists to easily classify their own sequences of interest, available at [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff).
 
@@
 Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv).
-The pipeline in Figure \@ref(effectorcollection) outlines the steps used.
+The pipeline in `r fig_caption("effectorcollection", display="cite")` outlines the steps used.
 We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd).
@@
 The longest sequence  for bacteria, fungi, and oomycete are 2574, 4034, and 934, respectively.
-Encoded sequences were split into taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd)
+Encoded sequences were split into random taxon specific training, test and validation sets at a 60%, 20%, 20% split respectively as described in code at [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0008_split_and_encode.Rmd).
 
@@
 Hyperparameter scans were performed using random search, a hyperparameter optimization method where each hyperparameter setting is randomly sampled from a distribution of possible hyperparameter values [@Bergstra2012].
+Hyperparameters to vary were selected as the ones generally known and expected to have strongest effect.
 We used RandomSearchCV(), the implementation of random search in `scikit-learn 0.19.2` [@scikitlearn] together with KerasClassifier() which is an implementation of the scikit-learn classifier API for keras.
@@
 Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).
+Five-fold cross validation was used in all instances.
 
@@
 ```{r, bacefftable, echo=FALSE, message = FALSE}
-apa_table(
-  readr::read_tsv("tables/bacterial_effector_count.tab"),
+  readr::read_tsv("tables/bacterial_effector_count.tab") %>% 
+  knitr::kable( "latex",
   digits = 0,
   align = 'c',
-  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base',
-  placement = "h"
-)
+  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
+) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header","hold_position"), font_size = 7)
 ```
@@
 ```{r, fungiefftable}
-apa_table(
-  readr::read_tsv("tables/fungal_effector_count.tab"),
+  readr::read_tsv("tables/fungal_effector_count.tab") %>% 
+  knitr::kable( "latex",
   digits = 0,
@@
   caption = 'Fungal species contributing to the fungal effector sequence set retrieved from PHI-Base'
-)
+) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header","hold_position"), font_size = 7)
 ```
@@
 ```{r, oomefftable}
-apa_table(
-  readr::read_tsv("tables/oomycete_effector_count.tab"),
+  readr::read_tsv("tables/oomycete_effector_count.tab") %>% 
+  knitr::kable( "latex",
   digits = 0,
@@
   caption = 'Oomycete species contributing to the oomycete effector sequence set retrieved from PHI-Base'
-)
+) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header","hold_position"), font_size = 7)
 ```
@@
 ```{r, possiblecombination}
-apa_table(
-  readr::read_tsv("tables/possible_combination_each_model.tab"),
+
+  readr::read_tsv("tables/possible_combination_each_model.tab")%>% 
+  knitr::kable( "latex",
   digits = 0,
@@
   caption = 'List of possible combination of hyperparamaters setting for each model'
-)
-```
-
+) %>%
+  kableExtra::kable_styling(latex_options = c("repeat_header","hold_position"), font_size = 7)
+```
+\newpage
 ```{r, fixedhyperparam}
@@
   ) %>%
-  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
-```
-
+  kableExtra::kable_styling(latex_options = c("repeat_header","hold_position"), font_size = 7)
+```
+\newpage
 ```{r, tunedhyperparam}
@@
   ) %>%
-  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
+  kableExtra::kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 7)
 ```
